---
title: "EDA Adventure: Meet the Employees"
author: "Add your gourps name here"
output: html_document
date: "2025-09-25"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- When you finish, please send it tou our emails: fmig@fct.unl.pt and m.norouzirad@fct.unl.pt -->

## Welcome to the Employee Data Adventure

Today, you are data detectives️

You have just received a secret dataset from a mysterious company.

Your job: uncover the hidden stories about the people who work there.

```{r}
# Simple code
url <- "https://raw.githubusercontent.com/mnrzrad/SIE/main/data/EmployeeData.csv"
df <- read.csv(url, stringsAsFactors = FALSE)

# Advanced R - using package "readr" (tidyverse)
# library(readr)
# df <- read_csv(url)
```

### Level 1: Is data here?

The dataset has just arrived in your R session.

-   Open it and check if it’s ready.

-   Peek inside the data — do you see rows and columns?

-   Explore it a little: what stands out first?

```{r}

```

### Level 2: Meet the data

Imagine walking into the company lobby for the first time and you are
meeting a new group of people. Take a look at the dataset as if you are
greeting them.

-   Notice how many people are in the room. Who is here?

-   Look at the kinds of information each person carries with them.
    Which information? What type of Information?

```{r}


```

**Dataset columns:**

-   **Employee_ID** – unique identifier for each employee •

-   **Name** – employee name

-   **Age** – employee age

-   **Gender** – Male/Female

-   **Department** – employee department (e.g., HR, Sales, Engineering,
    Finance)

-   **Job_Title** – job title (e.g., Analyst, Engineer, Executive,
    Intern)

-   **Experience_Years** – years of professional experience

-   **Education_Level** – highest education attained (Bachelor, Master,
    etc.)

-   **Location** – city of employment (Austin, Seattle, New York, etc.)

-   **Salary** – annual salary in USD

## Level 3: Discover the groups

Walk among the employees. Notice patterns.

Some employees belong to the same department. Some have the same job
title.

-   Look for groups and see which ones are large or small.

-   Which departments have the most people? Which have few?

-   Are some job titles more common than others?

-   Can you describe groups of employees who are similar?

```{r}


```

## Level 4: Follow the money

Salaries are hidden treasures.

-   Who earns a lot? Who earns less?

-   Do some departments pay more than others?

-   Do some job titles or education levels get higher salaries?

```{r}


```

## Level 5: Find hidden patterns

Some secrets are not obvious at first.

-   Does age relate to salary?

-   Does gender make a difference in pay?

-   Are there other interesting patterns you can find?

-   Can you visualize patterns?

```{r}


```

## Final Step: Your own exploration

Now the adventure is yours.

-   Ask your own questions.

-   Explore freely.

-   Find one story, one pattern, or one surprise in the data and share
    it with the class.

```{r}


```
