---
engine: knitr
filters:
  - webr
---

```{r setup, include=FALSE}
required_packages <- c("ggplot2", "dplyr", "factoextra", "corrplot", "gridExtra",
                       "GPArotation", "tibble")

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

# Principal Component Analysis

*Principal Component Analysis* (PCA) is a technique used to replace a set of correlated variables by a smaller set of new, uncorrelated variables — the *principal components* — that retain as much of the original variability as possible.

Where clustering looks for structure among the **rows** of a data set (which clients resemble each other?), PCA looks for structure among the **columns** (which variables carry the same information?). The two questions are complementary, and in practice PCA is very often the step that comes *before* clustering or classification.

PCA was introduced by @pearson1901 as a problem of fitting lines and planes to a cloud of points, and developed in its modern statistical form by @hotelling1933. A comprehensive treatment is given by @jolliffe2002.

## Why Reduce Dimension?

A retail bank stores dozens — often hundreds — of numeric attributes per client: balances across several account types, transaction counts by channel, credit exposure, tenure, product holdings, contact history. An insurer stores a similar volume per policy. Working with all of them at once creates four practical difficulties.

**Redundancy**

Many variables measure the same underlying thing. A client's income, current-account balance, savings balance, investment holdings and credit limit are all partial views of a single unobserved quantity we might call *affluence*. Keeping five columns to describe one construct adds noise, not information.

**Multicollinearity**

Strongly correlated predictors destabilise regression models: coefficients become large, sign-flipped and highly sensitive to small changes in the sample. Because principal components are *uncorrelated by construction*, they offer one way out of this problem.

**Visualisation**

We can look at a scatterplot of two variables, and with effort at a matrix of pairwise plots for five or six. We cannot look at a thirteen-dimensional cloud. PCA gives us the best possible two-dimensional "photograph" of that cloud, in a precise sense defined below.

**The cost of dimension**

Distances, densities and neighbourhoods behave badly in high dimension — the phenomenon usually called the *curse of dimensionality*. Since $k$-means, $k$-medoids and $k$-nearest-neighbours all rest on distances, reducing dimension first often improves both their speed and their results.

:::{.callout-note}
PCA does **not** select a subset of the original variables. It builds *new* variables, each one a weighted combination of all the originals. If your goal is to discard columns and keep the rest (for example, because measuring each variable is expensive), PCA is the wrong tool — you want variable selection instead.
:::

### Applications in Banking and Insurance

**Client segmentation**

Reducing a wide client table to two or three interpretable axes — affluence, digital engagement, credit strain — and then clustering on those axes rather than on the raw variables.

**Credit and behavioural scoring**

Compressing correlated bureau and transaction variables into orthogonal components before fitting a scorecard, so that coefficient estimates remain stable.

**Yield-curve and market-risk modelling**

The classic financial application: applied to daily changes in interest rates across maturities, PCA almost always returns three components interpretable as *level*, *slope* and *curvature*, which together explain over 95% of yield-curve movement. Risk systems hedge these three factors instead of every maturity separately.

**Insurance portfolio analysis**

Summarising policyholder and claims characteristics (exposure, frequency, severity, coverage limits, bonus-malus level) into a small number of risk dimensions for tariff analysis or reinsurance decisions.

**Fraud and anomaly detection**

An observation that is poorly reconstructed from the first few components — that is, one lying far from the low-dimensional subspace where ordinary clients live — is a natural candidate for investigation.

**Questionnaire and survey analysis**

Identifying the latent dimensions behind a battery of customer-satisfaction items, which are typically far more numerous than the concepts they measure.

:::{.callout-note appearance="minimal"}
**Packages and commands.** No code in this section. The chapter uses **stats** for the analysis itself (`prcomp`, later `varimax` and `promax`), **dplyr** for the data preparation, **factoextra** and **corrplot** for the standard PCA figures, **gridExtra** to place figures side by side, and **GPArotation** for the rotation criteria that **stats** does not provide. All are installed and loaded by the setup chunk at the top of the chapter.
:::

## The Data: A Retail Bank Client Portfolio

Throughout this chapter we work with a simulated portfolio of $600$ retail bank clients. The data are generated rather than downloaded so that every result in this chapter is exactly reproducible in your browser, but the variables, their units and their correlation structure are typical of a real retail banking table.

| Variable | Type | Description | Unit |
|:------------------|:---------|:-----------------------------------|:---------|
| `client_id` | id | Anonymised client identifier | — |
| `age` | numeric | Age of the client | years |
| `tenure` | numeric | Time as a client of the bank | years |
| `income` | numeric | Declared annual net income | € |
| `balance_current` | numeric | Average balance, current account | € |
| `balance_savings` | numeric | Average balance, savings products | € |
| `investments` | numeric | Value of investment holdings | € |
| `credit_limit` | numeric | Total credit-card limit granted | € |
| `utilization` | numeric | Share of the credit limit in use | % |
| `loan_balance` | numeric | Outstanding loan and mortgage debt | € |
| `card_tx` | numeric | Card transactions per month | count |
| `mobile_logins` | numeric | Mobile-app logins per month | count |
| `branch_visits` | numeric | Branch visits per year | count |
| `missed_payments` | numeric | Missed payments in the last 24 months | count |
| `segment` | factor | Commercial segment: Mass / Affluent / Private | — |
| `region` | factor | Region of residence | — |
| `churn` | factor | Client closed the relationship in the last year | — |

The code below creates the data set. Run it once: every interactive box in this chapter uses the object `bank` that it leaves behind.

```{r}
#| label: generate-bank
#| echo: false
#| message: false
#| warning: false

set.seed(2025)
n <- 600

W <- rnorm(n)                     # latent affluence
A <- rnorm(n)                     # latent life stage
D <- 0.80 * rnorm(n) - 0.45 * A   # latent digital propensity
R <- rnorm(n)                     # latent credit risk

age    <- round(pmin(pmax(rnorm(n, 45 + 12 * A + 2 * W, 4), 19), 84))
tenure <- round(pmin(pmax(rnorm(n, (age - 18) * 0.32 + 1.5 * A, 3), 0.5), 40), 1)

income          <- round(exp(rnorm(n, log(26000) + 0.45 * W, 0.30)), -2)
balance_current <- round(exp(rnorm(n, log(2800)  + 0.55 * W + 0.10 * D, 0.60)), -1)
balance_savings <- round(exp(rnorm(n, log(11000) + 0.90 * W + 0.25 * A, 0.70)), -1)
investments     <- round(exp(rnorm(n, log(4500)  + 1.25 * W + 0.20 * A, 1.00)), -1)
credit_limit    <- round(exp(rnorm(n, log(3000)  + 0.60 * W, 0.40)), -2)
utilization     <- round(pmin(pmax(rnorm(n, 34 + 17 * R - 6 * W, 9), 0), 100), 1)
loan_balance    <- round(exp(rnorm(n, log(14000) + 0.55 * R + 0.40 * W, 0.80)), -2)
card_tx         <- rpois(n, exp(log(24) + 0.45 * D + 0.15 * W))
mobile_logins   <- rpois(n, exp(log(16) + 0.60 * D))
branch_visits   <- rpois(n, exp(log(5)  - 0.70 * D + 0.15 * A))
missed_payments <- pmin(rpois(n, exp(log(0.6) + 0.80 * R - 0.30 * W)), 12)

wealth_score <- 0.5 * scale(log(income)) + 0.5 * scale(log(investments + 1))
segment <- cut(wealth_score,
               breaks = quantile(wealth_score, c(0, .60, .90, 1)),
               labels = c("Mass", "Affluent", "Private"),
               include.lowest = TRUE)
region  <- factor(sample(c("Norte", "Centro", "Lisboa", "Sul"), n,
                         TRUE, c(.28, .22, .35, .15)))
eta     <- -1.6 - 0.9 * D + 0.6 * R
churn   <- factor(ifelse(runif(n) < 1 / (1 + exp(-eta)), "Yes", "No"),
                  levels = c("No", "Yes"))

bank <- data.frame(client_id = sprintf("C%04d", 1:n),
                   age, tenure, income, balance_current, balance_savings,
                   investments, credit_limit, utilization, loan_balance,
                   card_tx, mobile_logins, branch_visits, missed_payments,
                   segment, region, churn)

num_vars <- c("age", "tenure", "income", "balance_current", "balance_savings",
              "investments", "credit_limit", "utilization", "loan_balance",
              "card_tx", "mobile_logins", "branch_visits", "missed_payments")
```

```{webr-r}
#| label: generate-bank-webr
#| autorun: true

set.seed(2025)
n <- 600

W <- rnorm(n)                     # latent affluence
A <- rnorm(n)                     # latent life stage
D <- 0.80 * rnorm(n) - 0.45 * A   # latent digital propensity
R <- rnorm(n)                     # latent credit risk

age    <- round(pmin(pmax(rnorm(n, 45 + 12 * A + 2 * W, 4), 19), 84))
tenure <- round(pmin(pmax(rnorm(n, (age - 18) * 0.32 + 1.5 * A, 3), 0.5), 40), 1)

income          <- round(exp(rnorm(n, log(26000) + 0.45 * W, 0.30)), -2)
balance_current <- round(exp(rnorm(n, log(2800)  + 0.55 * W + 0.10 * D, 0.60)), -1)
balance_savings <- round(exp(rnorm(n, log(11000) + 0.90 * W + 0.25 * A, 0.70)), -1)
investments     <- round(exp(rnorm(n, log(4500)  + 1.25 * W + 0.20 * A, 1.00)), -1)
credit_limit    <- round(exp(rnorm(n, log(3000)  + 0.60 * W, 0.40)), -2)
utilization     <- round(pmin(pmax(rnorm(n, 34 + 17 * R - 6 * W, 9), 0), 100), 1)
loan_balance    <- round(exp(rnorm(n, log(14000) + 0.55 * R + 0.40 * W, 0.80)), -2)
card_tx         <- rpois(n, exp(log(24) + 0.45 * D + 0.15 * W))
mobile_logins   <- rpois(n, exp(log(16) + 0.60 * D))
branch_visits   <- rpois(n, exp(log(5)  - 0.70 * D + 0.15 * A))
missed_payments <- pmin(rpois(n, exp(log(0.6) + 0.80 * R - 0.30 * W)), 12)

wealth_score <- 0.5 * scale(log(income)) + 0.5 * scale(log(investments + 1))
segment <- cut(wealth_score,
               breaks = quantile(wealth_score, c(0, .60, .90, 1)),
               labels = c("Mass", "Affluent", "Private"),
               include.lowest = TRUE)
region  <- factor(sample(c("Norte", "Centro", "Lisboa", "Sul"), n,
                         TRUE, c(.28, .22, .35, .15)))
eta     <- -1.6 - 0.9 * D + 0.6 * R
churn   <- factor(ifelse(runif(n) < 1 / (1 + exp(-eta)), "Yes", "No"),
                  levels = c("No", "Yes"))

bank <- data.frame(client_id = sprintf("C%04d", 1:n),
                   age, tenure, income, balance_current, balance_savings,
                   investments, credit_limit, utilization, loan_balance,
                   card_tx, mobile_logins, branch_visits, missed_payments,
                   segment, region, churn)

num_vars <- c("age", "tenure", "income", "balance_current", "balance_savings",
              "investments", "credit_limit", "utilization", "loan_balance",
              "card_tx", "mobile_logins", "branch_visits", "missed_payments")

str(bank)
```

Before any multivariate method, look at the marginal distributions. This is exactly the exploratory work of the EDA chapter, and it determines the choices we make next.

```{r}
#| label: bank-summary
summary(bank[, c("income", "balance_savings", "investments",
                 "utilization", "mobile_logins", "missed_payments")])
```

### Skewness and the Log Transform

The monetary variables are strongly right-skewed, as monetary variables almost always are: most clients hold modest amounts and a small number hold very large ones.

```{r}
#| label: skewness
skewness <- function(x) mean((x - mean(x))^3) / sd(x)^3
round(sapply(bank[num_vars], skewness), 2)
```

The savings balance has a skewness of $15.1$ and investments of $7.3$. PCA is a *variance*-based method, and variance is not robust: a handful of private-banking clients with balances two orders of magnitude above the median would, on the raw scale, define the first component almost by themselves. Taking logarithms of the monetary variables brings their skewness essentially to zero:

```{r}
#| label: build-X
#| message: false
library(dplyr)

X <- bank %>%
  transmute(age, tenure,
            log_income  = log(income),
            log_current = log(balance_current),
            log_savings = log(balance_savings),
            log_invest  = log(investments),
            log_limit   = log(credit_limit),
            utilization,
            log_loan    = log(loan_balance),
            card_tx,
            mobile      = mobile_logins,
            branch      = branch_visits,
            missed      = missed_payments)

round(sapply(X, skewness), 2)
```

The same matrix is built in your browser session by the helper below. Every interactive box from here on begins with `X <- make_X()`, so you can run them in any order.

```{webr-r}
#| label: make-X-webr
#| autorun: true

make_X <- function(d = bank) {
  data.frame(age         = d$age,
             tenure      = d$tenure,
             log_income  = log(d$income),
             log_current = log(d$balance_current),
             log_savings = log(d$balance_savings),
             log_invest  = log(d$investments),
             log_limit   = log(d$credit_limit),
             utilization = d$utilization,
             log_loan    = log(d$loan_balance),
             card_tx     = d$card_tx,
             mobile      = d$mobile_logins,
             branch      = d$branch_visits,
             missed      = d$missed_payments)
}

X <- make_X()
str(X)
```

:::{.callout-tip}
Log transformation of monetary variables before PCA is standard practice in banking and insurance analytics. It has three effects: it removes the dominance of a few very large clients, it turns *multiplicative* relationships (a client with twice the income tends to hold twice the savings) into *additive* ones, which is what linear methods can capture, and it makes the resulting components easier to read as differences in order of magnitude rather than in euros.

Beware of zeros: $\log(0)$ is undefined. Common remedies are $\log(x+1)$, or treating "holds no investments" as a separate binary variable.
:::

The count variables `card_tx`, `mobile` and `branch` keep some skewness, which is normal for counts and not damaging at this level. We leave them untransformed so that their scale stays interpretable.

### The Correlation Matrix

PCA is, in the end, a way of reading a correlation matrix. It is worth looking at that matrix directly first.

```{r}
#| label: corrplot
#| fig-cap: "Correlation matrix of the transformed bank variables. Blocks of mutually correlated variables are exactly what PCA will summarise."
#| fig-width: 7
#| fig-height: 6
#| message: false
library(corrplot)
corrplot(cor(X), method = "color", type = "upper",
         order = "hclust", tl.col = "black", tl.srt = 45,
         addCoef.col = "grey30", number.cex = 0.55)
```

Three blocks stand out:

- the monetary variables (`log_income`, `log_savings`, `log_invest`, `log_limit`, `log_current`) move together;
- the channel variables (`card_tx`, `mobile`) move together and both move *against* `branch`, `age` and `tenure`;
- `utilization`, `log_loan` and `missed` form a third, smaller block.

If the correlation matrix were the identity — all variables mutually uncorrelated — PCA would have nothing to do: every component would explain exactly $1/p$ of the variance and no reduction would be possible. **PCA is useful precisely to the extent that the variables are correlated.**

```{webr-r}
#| label: explore-cor
#| autorun: false

# Try it: which pairs are most strongly correlated?
Xsub <- data.frame(age = bank$age, tenure = bank$tenure,
                   log_income  = log(bank$income),
                   log_savings = log(bank$balance_savings),
                   utilization = bank$utilization,
                   mobile      = bank$mobile_logins,
                   branch      = bank$branch_visits)

round(cor(Xsub), 2)
```

:::{.callout-note appearance="minimal"}
**Packages and commands.** Simulation with `set.seed`, `rnorm`, `rpois`, `runif` and `sample`; marginal distributions with `summary` and `sapply`; the log transform with **dplyr**'s `transmute`; the correlation matrix with `cor` and **corrplot**'s `corrplot(method = "color", order = "hclust")`, whose `order` argument is what makes the blocks visible.
:::

## Pre-PCA Diagnostic Tests {#sec-pre-pca-tests}

The previous section ended on a qualitative statement: PCA is useful to the extent that the variables are correlated. Two classical diagnostics turn that statement into numbers, and they are routinely reported before a PCA:

- **Bartlett's test of sphericity** asks whether there is *any* correlation to summarise;
- the **Kaiser–Meyer–Olkin (KMO) measure** asks whether that correlation is *shared* across many variables, which is what a small number of components can capture, or confined to isolated pairs.

Both come from the factor-analysis tradition. They do not decide whether PCA can be computed — it always can — but whether it is likely to produce a useful reduction.

### Bartlett's Test of Sphericity

Bartlett's test [@bartlett1950] tests the hypothesis that the population correlation matrix is the identity:
$$
H_0: \mathbf{R} = \mathbf{I}_p
\qquad \text{versus} \qquad
H_1: \mathbf{R} \neq \mathbf{I}_p .
$$
Under $H_0$ the variables are uncorrelated and, in the language of the previous section, PCA has nothing to do. The test statistic is built from the determinant of the sample correlation matrix $\hat{\mathbf{R}}$:
$$
\chi^2 = -\left(n - 1 - \frac{2p + 5}{6}\right) \ln |\hat{\mathbf{R}}| ,
$$
which, under $H_0$ and multivariate normality, is approximately $\chi^2$-distributed with $p(p-1)/2$ degrees of freedom.

The determinant is the natural quantity to look at. It equals the product of the eigenvalues of $\hat{\mathbf{R}}$, so $|\hat{\mathbf{R}}| = 1$ when the variables are uncorrelated and $|\hat{\mathbf{R}}| \to 0$ as they become redundant — that is, as some eigenvalues approach zero. A small determinant means a large statistic.

```{r}
#| label: bartlett
bartlett_sphericity <- function(X) {
  R    <- cor(X)
  n    <- nrow(X)
  p    <- ncol(X)
  stat <- -(n - 1 - (2 * p + 5) / 6) * log(det(R))
  df   <- p * (p - 1) / 2
  c(det = det(R), chisq = stat, df = df,
    p.value = pchisq(stat, df, lower.tail = FALSE))
}

signif(bartlett_sphericity(X), 4)
```

The determinant is $0.0007$, the statistic is $\chi^2 = 4316$ on $78$ degrees of freedom, and the $p$-value is zero to machine precision. The hypothesis of no correlation is rejected overwhelmingly.

That conclusion is less informative than it looks. The test has enormous power at realistic sample sizes, so it rejects for correlations far too weak to be worth summarising. Compare a matrix of pure noise of the same shape with the first $30$ clients of our portfolio:

```{r}
#| label: bartlett-compare
set.seed(1)
noise <- matrix(rnorm(600 * 13), nrow = 600, ncol = 13)
signif(bartlett_sphericity(noise), 4)       # 600 x 13, no structure
signif(bartlett_sphericity(X[1:30, ]), 4)   # only 30 clients
```

On noise the test correctly fails to reject ($\chi^2 = 76.7$, $p = 0.52$); on just $30$ clients it already rejects with $p \approx 10^{-21}$. Passing Bartlett's test is therefore a **minimum requirement**, not a recommendation: if it does *not* reject, stop — there is nothing for PCA to find. If it does, move on to KMO.

:::{.callout-warning}
`stats::bartlett.test()` is **not** this test. It is Bartlett's test for *homogeneity of variances* across groups, by the same author, and applying it to a data frame returns a meaningless result without an error. Use the function above or `psych::cortest.bartlett()`.

The $\chi^2$ approximation also assumes multivariate normality. Our count variables and the bounded `utilization` are clearly not normal, so treat the $p$-value as approximate — which, given how far it sits from any threshold, changes nothing here.
:::

### The Kaiser–Meyer–Olkin Measure of Sampling Adequacy

The KMO measure [@kaiser1970; @kaiser1974] compares the ordinary correlations with the **partial correlations** — the correlation between two variables after the linear effect of all the others has been removed. The partial correlations are read off the inverse of the correlation matrix, $\mathbf{Q} = \mathbf{R}^{-1}$:
$$
a_{ij} = -\frac{q_{ij}}{\sqrt{q_{ii}\, q_{jj}}} .
$$

The logic is this. If variables $i$ and $j$ are correlated *because* both reflect a common underlying dimension that other variables also reflect, then controlling for those other variables removes most of the link and $a_{ij}$ is small. If instead the two are linked only to each other, the partial correlation stays as large as the raw one. Components summarise the first kind of correlation, not the second.

The overall index compares the two sums of squares over all pairs $i \neq j$:
$$
\text{KMO} = \frac{\sum_{i}\sum_{j \neq i} r_{ij}^2}{\sum_{i}\sum_{j \neq i} r_{ij}^2 + \sum_{i}\sum_{j \neq i} a_{ij}^2} ,
$$
and the **measure of sampling adequacy** of a single variable restricts the sums to its own column:
$$
\text{MSA}_j = \frac{\sum_{i \neq j} r_{ij}^2}{\sum_{i \neq j} r_{ij}^2 + \sum_{i \neq j} a_{ij}^2} .
$$
Both lie between $0$ and $1$. Values near $1$ mean the partial correlations are negligible compared with the raw ones; a value of $0.5$ is what uncorrelated noise produces. Kaiser's own labels for the scale are still the ones in use:

| KMO | Kaiser's verdict |
|:-----------|:-----------------|
| $\geq 0.90$ | marvellous |
| $0.80$ – $0.89$ | meritorious |
| $0.70$ – $0.79$ | middling |
| $0.60$ – $0.69$ | mediocre |
| $0.50$ – $0.59$ | miserable |
| $< 0.50$ | unacceptable |

These labels are Kaiser's verbal conventions from factor analysis, not inferential cut-offs: no sampling distribution stands behind them, and a *middling* value does not stop PCA from finding clear components, as the rest of this chapter shows.

```{r}
#| label: kmo
kmo <- function(X) {
  R <- cor(X)
  Q <- solve(R)
  A <- -Q / sqrt(diag(Q) %o% diag(Q))   # partial correlations
  diag(R) <- 0
  diag(A) <- 0
  list(overall = sum(R^2) / (sum(R^2) + sum(A^2)),
       msa     = colSums(R^2) / (colSums(R^2) + colSums(A^2)))
}

kmo_X <- kmo(X)
round(kmo_X$overall, 3)
round(sort(kmo_X$msa), 3)
round(kmo(noise)$overall, 3)
```

The overall KMO is $0.784$: *middling*, close to *meritorious*, and comfortably adequate for PCA. Pure noise gives $0.491$, as expected.

The per-variable values are more instructive than the overall one. The five monetary variables all score between $0.89$ and $0.93$: their raw correlations with one another lie between $0.51$ and $0.71$, but their partial correlations only between $0.06$ and $0.31$, because the block explains each pair jointly. The lowest scores — `utilization` ($0.61$), `mobile` ($0.64$), `tenure` ($0.66$), `card_tx` ($0.66$), `age` ($0.67$) and `missed` ($0.68$) — belong to variables whose correlation is concentrated in one or two partners:

```{r}
#| label: partial-cor
Rm <- cor(X)
Qm <- solve(Rm)
Am <- -Qm / sqrt(diag(Qm) %o% diag(Qm))   # partial correlations

pairs <- rbind(c("age", "tenure"), c("card_tx", "mobile"),
               c("utilization", "missed"), c("utilization", "log_loan"),
               c("log_savings", "log_invest"))
data.frame(pair    = paste(pairs[, 1], "-", pairs[, 2]),
           raw     = round(Rm[pairs], 2),
           partial = round(Am[pairs], 2))
```

`age` and `tenure` are correlated at $0.86$, and after removing the other eleven variables they are still correlated at $0.81$: older clients have simply been clients for longer, and no other variable explains that. The same holds for `card_tx` and `mobile` ($0.82 \to 0.79$) and for the credit-strain trio. The monetary pair, by contrast, falls from $0.65$ to $0.20$: most of what links savings and investments is the affluence that every other monetary variable also measures.

A low MSA does not make a variable useless to PCA — these same variables will define the digital-engagement and credit-strain components later in the chapter. It says that their correlation is *local*, so that it takes a component to describe a small group of variables rather than a large one. The usual rule is to consider dropping variables with $\text{MSA}_j < 0.5$, one at a time and starting from the lowest, recomputing after each removal. None of ours comes close, so all thirteen are kept.

:::{.callout-important}
## When the diagnostics cannot be computed

KMO needs $\mathbf{R}^{-1}$ and Bartlett needs $\ln|\mathbf{R}|$, so both fail when $\mathbf{R}$ is singular: when $p \geq n$, or when one variable is an exact linear combination of others — a total alongside its parts, or a full set of dummy variables. `solve` then stops with an error and `det` returns zero. Remove the redundancy first. PCA itself still runs in these cases, and is often exactly the right tool for them; it is only the diagnostics that break.
:::

```{webr-r}
#| label: pre-pca-tests-webr
#| autorun: false

X <- make_X()

bartlett_sphericity <- function(X) {
  R    <- cor(X)
  n    <- nrow(X); p <- ncol(X)
  stat <- -(n - 1 - (2 * p + 5) / 6) * log(det(R))
  df   <- p * (p - 1) / 2
  c(chisq = stat, df = df, p.value = pchisq(stat, df, lower.tail = FALSE))
}

kmo <- function(X) {
  R <- cor(X); Q <- solve(R)
  A <- -Q / sqrt(diag(Q) %o% diag(Q))
  diag(R) <- 0; diag(A) <- 0
  list(overall = sum(R^2) / (sum(R^2) + sum(A^2)),
       msa     = colSums(R^2) / (colSums(R^2) + colSums(A^2)))
}

bartlett_sphericity(X)
round(sort(kmo(X)$msa), 3)

# Try it: drop `tenure`. Which variable's MSA changes most, and why?
# Then try keeping only the five monetary variables.
round(kmo(X[, setdiff(names(X), "tenure")])$overall, 3)
```

:::{.callout-note appearance="minimal"}
**Packages and commands.** Both diagnostics need only base R: `cor`, `det` and `pchisq` for Bartlett's statistic, and `solve` with the outer product `%o%` for the partial correlations behind KMO. The **psych** package wraps them as `cortest.bartlett(R, n = )` and `KMO(X)`. Do not confuse either with `stats::bartlett.test`, which tests equality of variances.
:::

## Notation, Centering and Scaling

Let $\mathbf{X}$ be an $n \times p$ data matrix: $n$ clients in rows, $p$ variables in columns. Write $\mathbf{x}_i = (x_{i1}, \ldots, x_{ip})^{\top}$ for the record of client $i$, and $\bar{x}_j$ and $s_j$ for the sample mean and standard deviation of variable $j$.

### Centering

PCA always begins by *centering* each variable:
$$
\tilde{x}_{ij} = x_{ij} - \bar{x}_j .
$$
Centering moves the origin of the coordinate system to the centre of gravity of the cloud of points. It is not optional: the components are directions *through the mean*, and without centering the first component would mostly point from the origin towards the mean of the data, which tells us nothing about its shape.

### Scaling

*Scaling* is optional, and it is the single most consequential decision in a PCA. If we also divide by the standard deviation,
$$
z_{ij} = \frac{x_{ij} - \bar{x}_j}{s_j},
$$
then every variable has variance $1$, and PCA is performed on the **correlation matrix**. If we do not, PCA is performed on the **covariance matrix**, and variables with large numerical variance dominate.

Look at the raw variances in our data:

```{r}
#| label: raw-variances
round(apply(bank[, num_vars], 2, var))
```

The variance of `balance_savings` is of the order of $3.3 \times 10^{9}$ (euros squared); the variance of `missed_payments` is about $2$ (counts squared). These numbers are not comparable — they are not even in the same units. A PCA on the covariance matrix of these columns is a PCA of the savings balance and nothing else:

```{r}
#| label: pca-unscaled
pca_unscaled <- prcomp(bank[, num_vars], scale. = FALSE)
round(100 * pca_unscaled$sdev^2 / sum(pca_unscaled$sdev^2), 1)[1:4]
round(pca_unscaled$rotation[, 1:2], 3)
```

The first component absorbs $66.7\%$ of the variance and loads almost entirely on `balance_savings` and `investments`. Age, utilisation and every count variable have loadings indistinguishable from zero — not because they carry no information, but because they are measured in small numbers.

:::{.callout-warning}
When variables are measured in different units, **always** use the correlation matrix (`scale. = TRUE` in `prcomp`, or `cor = TRUE` in `princomp`). Otherwise the analysis reflects your choice of units rather than the structure of the data: converting income from euros to thousands of euros removes it from the first component altogether (its loading falls from $0.17$ to $0$), although nothing about the clients has changed.

The covariance matrix is appropriate only when all variables share the same unit and their differences in variability are themselves meaningful — for example, daily returns of a set of equities, or changes in yields across maturities.
:::

From here on, every PCA is on the correlation matrix of the transformed matrix `X`.

:::{.callout-note appearance="minimal"}
**Packages and commands.** `scale` centres and standardises a matrix, and `apply(X, 2, var)` shows why that matters here. `prcomp(X, scale. = TRUE)` runs the PCA on the correlation matrix and `scale. = FALSE` on the covariance matrix; the older `princomp` makes the same choice through `cor = TRUE`.
:::

## The Algebra of PCA

### The First Principal Component {#sec-first-pc}

Let $\mathbf{Z}$ be the centred and scaled data matrix, and consider a direction in variable space given by a unit vector $\mathbf{a} = (a_1, \ldots, a_p)^{\top}$, with $\mathbf{a}^{\top}\mathbf{a} = 1$. Projecting the data onto this direction produces a new variable
$$
y_i = \mathbf{a}^{\top}\mathbf{z}_i = a_1 z_{i1} + a_2 z_{i2} + \cdots + a_p z_{ip},
\qquad i = 1, \ldots, n ,
$$
a *linear combination* of the original variables. Its sample variance is
$$
\operatorname{Var}(y) = \mathbf{a}^{\top}\mathbf{S}\,\mathbf{a},
$$
where $\mathbf{S}$ is the sample covariance matrix of $\mathbf{Z}$ — which, because the variables are standardised, is the **correlation matrix** $\mathbf{R}$ of the original variables.

The first principal component is the direction that makes this variance as large as possible:
$$
\mathbf{a}_1 = \arg\max_{\mathbf{a}^{\top}\mathbf{a} = 1} \; \mathbf{a}^{\top}\mathbf{S}\,\mathbf{a}.
$$

The constraint $\mathbf{a}^{\top}\mathbf{a} = 1$ is essential: without it the variance could be made arbitrarily large simply by multiplying $\mathbf{a}$ by a large constant, which changes the scale of the new variable but not the direction it points in.

Introducing a Lagrange multiplier $\lambda$, we maximise
$$
\mathbf{a}^{\top}\mathbf{S}\mathbf{a} - \lambda(\mathbf{a}^{\top}\mathbf{a} - 1),
$$
and setting the derivative with respect to $\mathbf{a}$ to zero gives
$$
2\mathbf{S}\mathbf{a} - 2\lambda \mathbf{a} = \mathbf{0}
\qquad \Longleftrightarrow \qquad
\mathbf{S}\mathbf{a} = \lambda \mathbf{a}.
$$

This is the *eigenvalue equation*. The optimal direction $\mathbf{a}$ must be an **eigenvector** of $\mathbf{S}$, and its associated $\lambda$ is an **eigenvalue**. Moreover, pre-multiplying by $\mathbf{a}^{\top}$,
$$
\mathbf{a}^{\top}\mathbf{S}\mathbf{a} = \lambda\, \mathbf{a}^{\top}\mathbf{a} = \lambda ,
$$
so the variance captured by a component *is* its eigenvalue. To maximise variance we take the eigenvector belonging to the **largest** eigenvalue.

:::{.callout-note}
Two readings of the same result:

- **Maximum variance**: the first component is the direction along which the clients are most spread out.
- **Best fit**: the first component is also the line that minimises the sum of squared *perpendicular* distances from the points to the line. This is the geometry of @pearson1901, and it is not the same as the line fitted by ordinary least squares, which minimises *vertical* distances to a designated response variable.

Both descriptions define the same direction, which is why PCA can be presented either as variance maximisation or as low-dimensional approximation.
:::

### Subsequent Components

The second component solves the same problem with one extra requirement: it must be **uncorrelated** with the first. Since
$$
\operatorname{Cov}(\mathbf{a}_1^{\top}\mathbf{z},\; \mathbf{a}_2^{\top}\mathbf{z}) = \mathbf{a}_1^{\top}\mathbf{S}\mathbf{a}_2 = \lambda_1 \mathbf{a}_1^{\top}\mathbf{a}_2,
$$
requiring zero correlation is equivalent to requiring $\mathbf{a}_1^{\top}\mathbf{a}_2 = 0$: the directions are **orthogonal**. Continuing in this way, the $k$-th component is the direction of maximum remaining variance among those orthogonal to the first $k-1$.

Because $\mathbf{S}$ is symmetric and positive semi-definite, it admits the *spectral decomposition*
$$
\mathbf{S} = \mathbf{A}\,\boldsymbol{\Lambda}\,\mathbf{A}^{\top},
\qquad
\mathbf{A} = [\mathbf{a}_1, \ldots, \mathbf{a}_p],
\qquad
\boldsymbol{\Lambda} = \operatorname{diag}(\lambda_1, \ldots, \lambda_p),
$$
with $\lambda_1 \ge \lambda_2 \ge \cdots \ge \lambda_p \ge 0$ and $\mathbf{A}^{\top}\mathbf{A} = \mathbf{I}$. The columns of $\mathbf{A}$ are the $p$ principal directions, sorted by the variance they explain, and they form an orthonormal basis: PCA is a **rotation** of the coordinate system, nothing more, until we decide to discard some of the new axes.

### Total Variance is Preserved

Because the trace of a matrix is invariant under this rotation,
$$
\sum_{k=1}^{p} \lambda_k = \operatorname{tr}(\mathbf{S}) = \sum_{j=1}^{p} s_{jj}.
$$
On the correlation matrix every $s_{jj} = 1$, so
$$
\sum_{k=1}^{p} \lambda_k = p .
$$

With $p = 13$ variables, the eigenvalues of our correlation matrix must sum to $13$. This gives an immediate benchmark: a component with $\lambda_k = 1$ explains exactly as much variance as one original standardised variable. This is the basis of the Kaiser criterion in @sec-how-many.

The **proportion of variance explained** by component $k$ is therefore
$$
\text{PVE}_k = \frac{\lambda_k}{\sum_{m=1}^{p}\lambda_m} = \frac{\lambda_k}{p}
\quad \text{(correlation matrix)} .
$$

### Connection with the Singular Value Decomposition

In practice, software does not form $\mathbf{S}$ and diagonalise it; it applies the *singular value decomposition* to $\mathbf{Z}$ directly:
$$
\mathbf{Z} = \mathbf{U}\,\mathbf{D}\,\mathbf{V}^{\top},
$$
with $\mathbf{U}^{\top}\mathbf{U} = \mathbf{V}^{\top}\mathbf{V} = \mathbf{I}$ and $\mathbf{D}$ diagonal with non-negative entries $d_1 \ge \cdots \ge d_p$. Then
$$
\mathbf{S} = \frac{1}{n-1}\mathbf{Z}^{\top}\mathbf{Z}
= \frac{1}{n-1}\mathbf{V}\mathbf{D}^{2}\mathbf{V}^{\top},
$$
so $\mathbf{V} = \mathbf{A}$ (the loadings) and $\lambda_k = d_k^2/(n-1)$. The SVD route is numerically more stable, because squaring the data to form $\mathbf{S}$ squares the condition number as well. This is why `prcomp()` (SVD-based) is preferred in R over the older `princomp()` (eigen-based).

:::{.callout-note}
`prcomp()` divides by $n-1$; `princomp()` divides by $n$. The eigenvectors are identical, the eigenvalues differ by the factor $n/(n-1)$, and for $n = 600$ the difference is immaterial. Use `prcomp()`.
:::

### Doing It by Hand

Everything above can be verified in a few lines. First the eigen-decomposition of the correlation matrix:

```{r}
#| label: manual-eigen
R_mat <- cor(X)
e <- eigen(R_mat)

round(e$values, 3)          # the p eigenvalues
sum(e$values)               # must equal p = 13
```

Now compare with `prcomp()`:

```{r}
#| label: compare-prcomp
pca <- prcomp(X, scale. = TRUE)

max(abs(e$values - pca$sdev^2))                        # eigenvalues agree
max(abs(abs(e$vectors[, 1]) - abs(pca$rotation[, 1]))) # first direction agrees
```

Both differences are of the order of $10^{-14}$ — machine precision. And through the SVD:

```{r}
#| label: manual-svd
Z <- scale(X)                 # centred and scaled
s <- svd(Z)
max(abs(s$d / sqrt(nrow(X) - 1) - pca$sdev))
```

Identical again.

:::{.callout-warning}
**Signs are arbitrary.** If $\mathbf{a}$ is an eigenvector, so is $-\mathbf{a}$, with the same eigenvalue. Different software — and different versions of the same software — may return a component with all its signs flipped. This changes nothing about the analysis, but it does reverse the *verbal* interpretation ("high on PC3" becomes "low on PC3"), so always state the direction explicitly when reporting results. This is why the comparison above uses `abs()`.
:::

```{webr-r}
#| label: eigen-exercise
#| autorun: false

# Build the transformed matrix and reproduce the eigenvalues yourself
X <- make_X()

e <- eigen(cor(X))
round(e$values, 3)
sum(e$values)   # equals the number of variables
```

### A Two-Dimensional Picture of the Rotation

With thirteen variables the rotation cannot be drawn. With two it can, and the picture is worth more than the algebra. Take just `log_income` and `log_limit` — a client's declared income and the card limit the bank granted them, which are related for an obvious reason.

```{r}
#| label: rot-setup
pair  <- X[, c("log_income", "log_limit")]
pca2  <- prcomp(pair, scale. = TRUE)

# both loadings come back negative; flip so that PC1 points up and to the right
pca2$rotation <- -pca2$rotation
pca2$x        <- -pca2$x

cor(pair)[1, 2]
round(pca2$rotation, 4)
round(pca2$sdev^2, 4)
```

The correlation is $r = 0.713$. The left panel below shows the two standardised variables with the component directions drawn over the cloud; the right panel shows the *same clients* after the rotation, plotted against their scores.

```{r}
#| label: rotation-picture
#| echo: false
#| fig-cap: "Left: two correlated standardised variables, with the component axes drawn over the cloud. Right: the same 600 clients plotted on those axes. Nothing has moved relative to anything else — the page has been turned by 45 degrees. Both panels use a 1:1 aspect ratio, without which a rotation does not look like one."
#| fig-width: 10
#| fig-height: 5.4
#| message: false
#| warning: false
library(gridExtra)

Z2 <- as.data.frame(scale(pair)); names(Z2) <- c("z1", "z2")
L  <- pca2$sdev
V  <- pca2$rotation
lim <- 4

ax <- data.frame(comp = c("PC1", "PC2"),
                 x    = -2 * L * V[1, ], y    = -2 * L * V[2, ],
                 xend =  2 * L * V[1, ], yend =  2 * L * V[2, ])

arc <- data.frame(t = seq(0, pi / 4, length.out = 60))
arc$x <- 1.5 * cos(arc$t); arc$y <- 1.5 * sin(arc$t)

p_before <- ggplot(Z2, aes(z1, z2)) +
  geom_hline(yintercept = 0, colour = "grey75") +
  geom_vline(xintercept = 0, colour = "grey75") +
  geom_point(colour = "grey55", size = 1.1, alpha = 0.55) +
  geom_path(data = arc, aes(x, y), colour = "grey30", linewidth = 0.4) +
  annotate("text", x = 1.95, y = 0.75, label = "theta == 45*degree",
           parse = TRUE, size = 3.6, colour = "grey20") +
  annotate("text", x =  2.50, y = 2.5, label = "PC1",
           colour = "red3", fontface = 2, size = 4) +
  annotate("text", x = -1.55, y = 1.5, label = "PC2",
           colour = "steelblue", fontface = 2, size = 4) +
  geom_segment(data = ax, aes(x = x, y = y, xend = xend, yend = yend, colour = comp),
               linewidth = 1, arrow = arrow(length = unit(0.16, "cm"), ends = "both")) +
  scale_colour_manual(values = c(PC1 = "red3", PC2 = "steelblue"), guide = "none") +
  coord_equal(xlim = c(-lim, lim), ylim = c(-lim, lim)) +
  labs(title = "Before: the original standardised axes",
       subtitle = "correlated variables, component directions drawn over the cloud",
       x = "log_income (standardised)", y = "log_limit (standardised)") +
  theme_minimal() + theme(legend.position = "none")

p_after <- ggplot(as.data.frame(pca2$x), aes(PC1, PC2)) +
  geom_hline(yintercept = 0, colour = "red3", linewidth = 1) +
  geom_vline(xintercept = 0, colour = "steelblue", linewidth = 1) +
  geom_point(colour = "grey55", size = 1.1, alpha = 0.55) +
  coord_equal(xlim = c(-lim, lim), ylim = c(-lim, lim)) +
  labs(title = "After: rotated onto the component axes",
       subtitle = "same cloud, same distances, now uncorrelated",
       x = "PC1  (variance 1.713)", y = "PC2  (variance 0.287)") +
  theme_minimal() + theme(legend.position = "none")

grid.arrange(p_before, p_after, ncol = 2)
```

The elongated, tilted ellipse on the left becomes an ellipse aligned with the axes on the right. Three things are worth checking rather than believing:

```{r}
#| label: rot-checks
# 1. distances between clients are unchanged: a rotation is a rigid motion
d_before <- dist(scale(pair))
d_after  <- dist(pca2$x)
max(abs(d_before - d_after))

# 2. total variance is unchanged, only redistributed
c(before = sum(apply(scale(pair), 2, var)), after = sum(pca2$sdev^2))

# 3. the new variables are uncorrelated
round(cor(pca2$x)[1, 2], 12)
```

The variables started with variance $1$ each. After the rotation the same total of $2$ is split $1.713$ / $0.287$ — that is, $85.6\%$ of it now sits on a single axis. This is the whole of PCA in one line: **the rotation does not create or destroy variance, it concentrates it.**

<!-- ### Why the Angle is Exactly 45 Degrees

For two variables the rotation matrix is
$$
\mathbf{A} =
\begin{bmatrix} \cos\theta & -\sin\theta \\ \sin\theta & \cos\theta \end{bmatrix},
$$
and the angle that maximises the variance of the projection is
$$
\theta = \frac{1}{2}\arctan\!\left(\frac{2 s_{12}}{s_{11} - s_{22}}\right).
$$
When the variables are standardised, $s_{11} = s_{22} = 1$, the denominator vanishes, and the angle is $45^{\circ}$ whatever the value of $r$. The eigenvalues then have a closed form:
$$
\lambda_1 = 1 + r, \qquad \lambda_2 = 1 - r .
$$

```{r}
#| label: rot-closed-form
r <- cor(pair)[1, 2]
c(one_plus_r = 1 + r, one_minus_r = 1 - r)
round(pca2$sdev^2, 6)
```

:::{.callout-note}
For exactly two standardised variables the *direction* of the components carries no information at all — PC1 is always the bisector, PC2 always perpendicular to it. Only the eigenvalues say anything, and they say only what $r$ already said. PCA on two standardised variables is therefore never worth doing; it becomes informative from three variables upwards, where the direction of the first component genuinely depends on the pattern of correlations.
:::

The optimisation of @sec-first-pc is easy to see here: sweep a direction through every angle and record the variance of the data projected onto it. For standardised variables that variance is $1 + r\sin 2\theta$.

```{r}
#| label: angle-sweep-calc
R2  <- cor(pair)
ang <- seq(0, 180, by = 0.5)
va  <- sapply(ang, function(d) {
  a <- c(cos(d * pi / 180), sin(d * pi / 180))   # a unit vector at angle d
  as.numeric(t(a) %*% R2 %*% a)                  # t(a) S a
})

data.frame(angle = c(0, 45, 90, 135), variance = round(sapply(c(0, 45, 90, 135),
  function(d) { a <- c(cos(d * pi / 180), sin(d * pi / 180))
                as.numeric(t(a) %*% R2 %*% a) }), 3))
```

```{r}
#| label: angle-sweep-plot
#| echo: false
#| fig-cap: "Variance of the projected data as the projection direction sweeps through 180 degrees. PCA takes the maximum (PC1) and then the direction at right angles to it, which here is also the minimum (PC2)."
#| fig-width: 8
#| fig-height: 4.4
ggplot(data.frame(ang, va), aes(ang, va)) +
  geom_hline(yintercept = 1, linetype = "dotted", colour = "grey50") +
  geom_line(colour = "grey25", linewidth = 0.9) +
  geom_point(data = data.frame(ang = c(45, 135), va = c(1 + r, 1 - r)),
             aes(ang, va), colour = c("red3", "steelblue"), size = 3) +
  annotate("text", x = 45,  y = 1.80, parse = TRUE, size = 3.5, colour = "red3",
           label = "PC1: 45*degree*','~lambda[1]==1.713") +
  annotate("text", x = 135, y = 0.72, parse = TRUE, size = 3.5, colour = "steelblue",
           label = "PC2: 135*degree*','~lambda[2]==0.287") +
  scale_x_continuous(breaks = seq(0, 180, 45)) +
  labs(title = "Variance of the projection as the axis rotates",
       subtitle = "each angle is one candidate direction; PCA picks the maximum",
       x = "angle of the projection direction (degrees)",
       y = "variance of the projected data") +
  theme_minimal()
```

The dotted line at $1$ marks the variance of either original variable. Every direction between $0^{\circ}$ and $90^{\circ}$ does better than both of them, and the best does considerably better — which is the reason for rotating at all.

:::{.callout-tip}
In $p$ dimensions exactly the same thing happens: the cloud is a $p$-dimensional ellipsoid and PCA turns it until its longest axis lies along the first coordinate, its next longest along the second, and so on. What is lost above three dimensions is only our ability to look at it directly — hence the correlation circle and the biplot of @sec-visualising, which are two-dimensional windows onto that same rotation.
:::

```{webr-r}
#| label: rot-try
#| autorun: false

X <- make_X()

# Try another pair — for example age and tenure (r = 0.86),
# or card_tx and branch (r = -0.55). What changes, and what does not?
pair <- X[, c("age", "tenure")]
p2 <- prcomp(pair, scale. = TRUE)

cor(pair)[1, 2]
round(p2$sdev^2, 4)       # compare with 1 + r and 1 - r
round(p2$rotation, 4)     # still the bisector?

plot(scale(pair), pch = 19, col = "grey60", cex = 0.6, asp = 1,
     xlab = "variable 1 (standardised)", ylab = "variable 2 (standardised)")
abline(0,  1, col = "red3",     lwd = 2)
abline(0, -1, col = "steelblue", lwd = 2)
``` -->

:::{.callout-note appearance="minimal"}
**Packages and commands.** `cor` followed by `eigen` gives the eigenvalues (`$values`) and loadings (`$vectors`) by hand; `svd` gives the same through singular values (`$d`); `prcomp` does it in one call and is the one to use. `dist` checks that the rotation leaves every distance alone. The two-dimensional figure is **ggplot2** plus **gridExtra**'s `grid.arrange`, and depends on `coord_equal` — without a 1:1 aspect ratio a rotation does not look like one.
:::

## Scores, Loadings and Explained Variance

Three objects come out of a PCA, and confusing them is the most common source of error in reading PCA output.

| Object | Dimension | Meaning | In `prcomp()` |
|:-------------|:-----------|:--------------------------------|:-------------|
| **Loadings** | $p \times p$ | Weight of each *variable* in each component | `$rotation` |
| **Scores** | $n \times p$ | Coordinate of each *observation* on each component | `$x` |
| **Eigenvalues** | $p$ | Variance carried by each component | `$sdev^2` |

### Loadings

The loading $a_{jk}$ is the weight of variable $j$ in component $k$. The vector $\mathbf{a}_k$ has unit length, so the loadings of a component satisfy $\sum_j a_{jk}^2 = 1$; a variable's *contribution* to that component is therefore $100\,a_{jk}^2$ per cent.

### Scores

The score of client $i$ on component $k$ is
$$
y_{ik} = \mathbf{a}_k^{\top}\mathbf{z}_i = \sum_{j=1}^{p} a_{jk}\, z_{ij}.
$$
Scores are the new data: a client who was described by $13$ numbers is now described by $13$ new numbers, of which we intend to keep only the first few. By construction the scores have mean zero, variance $\lambda_k$, and zero correlation with each other.

### Fitting the PCA

```{r}
#| label: fit-pca
pca <- prcomp(X, scale. = TRUE)
summary(pca)
```

```{r}
#| label: eig-table
#| message: false
library(factoextra)
eig <- get_eigenvalue(pca)
round(eig, 3)
```

The eigenvalues are $4.05$, $2.93$, $1.76$ and $1.03$, explaining $31.2\%$, $22.6\%$, $13.5\%$ and $7.9\%$ of the total variance. The first three components together account for $67.2\%$ of the variability of the thirteen original variables; adding the fourth brings this to $75.1\%$.

:::{.callout-note}
It is normal — and healthy — for social and behavioural data not to be compressible into two components. A first component explaining $90\%$ of the variance usually means either that the variables are near-duplicates of one another, or that a scaling problem has gone unnoticed.
:::

### Reconstruction

Keeping only $k$ components and rotating back gives an approximation to the standardised data:
$$
\hat{\mathbf{Z}}_{(k)} = \mathbf{Y}_{(k)}\,\mathbf{A}_{(k)}^{\top},
$$
where $\mathbf{Y}_{(k)}$ holds the first $k$ score columns and $\mathbf{A}_{(k)}$ the first $k$ loading columns. The Eckart–Young theorem [@eckart1936] states that this is the **best possible rank-$k$ approximation** of $\mathbf{Z}$ in the least-squares sense: no other set of $k$ linear combinations reconstructs the data with smaller squared error.

```{r}
#| label: reconstruction
Z <- scale(X)
for (k in c(1, 2, 3, 4, 6, 8)) {
  Zhat <- pca$x[, 1:k, drop = FALSE] %*% t(pca$rotation[, 1:k, drop = FALSE])
  cat(sprintf("k = %d  mean squared error = %.3f   variance retained = %.1f%%\n",
              k, mean((Z - Zhat)^2), 100 * sum(pca$sdev[1:k]^2) / ncol(X)))
}
```

The error and the retained variance are two views of the same quantity: retaining $67.2\%$ of the variance with three components means leaving $32.8\%$ of it — a mean squared error of $0.327$ on standardised data — in the discarded directions.

```{webr-r}
#| label: pca-fit-webr
#| autorun: false

X <- make_X()

pca <- prcomp(X, scale. = TRUE)
summary(pca)

# scores of the first five clients on the first three components
round(pca$x[1:5, 1:3], 3)
```

:::{.callout-note appearance="minimal"}
**Packages and commands.** `prcomp` returns `$rotation` (loadings), `$x` (scores) and `$sdev` (square roots of the eigenvalues). `summary` prints the variance table; **factoextra**'s `get_eigenvalue` returns the same thing as a data frame. Reconstruction is plain matrix algebra: `pca$x[, 1:k] %*% t(pca$rotation[, 1:k])`.
:::

## How Many Components to Retain? {#sec-how-many}

There is no test that settles this question. The usual practice is to look at several criteria and take a decision that is defensible both statistically and substantively.

### The Scree Plot

Plot the eigenvalues against their index and look for an "elbow" — the point after which the curve flattens into a scree slope of small, roughly equal eigenvalues [@cattell1966]. Components before the elbow are taken as signal, those after as noise.

```{r}
#| label: scree
#| fig-cap: "Scree plot of the bank PCA. The curve bends after the third component; the fourth is borderline."
#| fig-width: 7
#| fig-height: 4.5
fviz_eig(pca, addlabels = TRUE, choice = "eigenvalue",
         barfill = "steelblue", barcolor = "steelblue") +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red3") +
  labs(title = "Scree plot", x = "Component", y = "Eigenvalue")
```

### Cumulative Proportion

Retain as many components as needed to reach a target share of the total variance — commonly $70\%$ or $80\%$. The threshold is a convention, not a result, and should be chosen with the use in mind: a visualisation may be satisfied with two components, a preprocessing step for modelling usually wants more.

```{r}
#| label: cumvar
#| fig-cap: "Cumulative percentage of variance explained. Three components pass 67%, four pass 75%."
#| fig-width: 7
#| fig-height: 4.5
cum <- data.frame(comp = 1:ncol(X), cum = cumsum(100 * pca$sdev^2 / ncol(X)))

ggplot(cum, aes(comp, cum)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2.5) +
  geom_hline(yintercept = c(70, 80), linetype = "dashed", colour = "grey40") +
  scale_x_continuous(breaks = 1:ncol(X)) +
  labs(title = "Cumulative variance explained",
       x = "Number of components", y = "Cumulative %") +
  theme_minimal()
```

### The Kaiser Criterion

On the correlation matrix, retain components with $\lambda_k > 1$ [@kaiser1960]. The logic is direct: a component explaining less than one standardised variable's worth of variance is not a summary of anything.

```{r}
#| label: kaiser
sum(pca$sdev^2 > 1)
round(pca$sdev^2, 3)
```

Four components pass. Note how narrowly: $\lambda_4 = 1.025$. A criterion that would drop a component if the fourth eigenvalue were $0.99$ and keep it at $1.01$ should not be applied mechanically.

### The Broken-Stick Model

If a stick of unit length is broken at random into $p$ pieces, the expected length of the $k$-th longest piece is
$$
b_k = \frac{1}{p}\sum_{m=k}^{p} \frac{1}{m}.
$$
Retain component $k$ if it explains more variance than $b_k$ — that is, more than would be expected if the total variance were divided among the $p$ directions purely at random. This is a stricter rule than Kaiser's, and generally a better-behaved one.

```{r}
#| label: broken-stick
p <- ncol(X)
bstick <- sapply(1:p, function(k) sum(1 / (k:p)) / p) * 100
obs    <- 100 * pca$sdev^2 / p

data.frame(component = 1:p,
           observed  = round(obs, 2),
           broken_stick = round(bstick, 2),
           retain = obs > bstick)[1:6, ]
```

Broken-stick retains **three** components; Kaiser retains **four**. The disagreement is informative rather than annoying: it tells us the fourth component sits on the boundary between structure and noise, and we should decide it on interpretive grounds.

:::{.callout-tip}
**Summary of retention criteria**

| Criterion | Rule | Retains here | Comment |
|:-----------|:------------------------|:-------:|:-----------------------------|
| Scree / elbow | Bend in the eigenvalue curve | 3 | Visual, subjective, widely used |
| Cumulative % | Reach 70–80% | 3–4 | Threshold is a convention |
| Kaiser | $\lambda_k > 1$ | 4 | Only for the correlation matrix; tends to over-retain |
| Broken stick | $\lambda_k/p > b_k$ | 3 | Stricter, generally reliable |
| Interpretability | Can you name it? | 3 | The criterion that decides ties — but see @sec-rotation |

When criteria disagree, prefer the solution you can explain to the business. A component nobody can name is rarely worth carrying into the next stage of the analysis.
:::

We retain **three** components for interpretation. Had the goal been prediction rather than explanation, keeping the fourth would have been defensible: a component that resists naming can still carry usable signal.

That is not quite the end of the question. @sec-rotation reopens it: a component that cannot be named *as it stands* may become nameable once the axes inside the retained subspace are rotated, and the fourth component of this portfolio is exactly such a case.

:::{.callout-note appearance="minimal"}
**Packages and commands.** **factoextra**'s `fviz_eig(choice = "eigenvalue")` draws the scree plot; `cumsum` and **ggplot2** draw the cumulative curve. The Kaiser count is `sum(pca$sdev^2 > 1)`, and the broken-stick values need no package at all — just `sapply` over the formula.
:::

## Visualising a PCA {#sec-visualising}

### The Variable Correlation Circle

Because the data are standardised, the correlation between variable $j$ and component $k$ is
$$
r(x_j, y_k) = a_{jk}\sqrt{\lambda_k},
$$
and these correlations are the coordinates used to plot variables on the *correlation circle*. Since a correlation cannot exceed $1$ in absolute value, every variable lies inside a circle of radius $1$.

```{r}
#| label: var-circle
#| fig-cap: "Variables on the plane of the first two components, coloured by quality of representation (cos2). Arrows close to the circle are well represented; small arrows are not."
#| fig-width: 7
#| fig-height: 6.5
fviz_pca_var(pca, axes = c(1, 2), col.var = "cos2",
             gradient.cols = c("grey70", "steelblue", "red3"),
             repel = TRUE) +
  labs(title = "Variables: PC1 vs PC2")
```

How to read this plot:

- **Direction**: variables pointing the same way are positively correlated; opposite directions mean negative correlation; a right angle between two well-represented arrows means near-zero correlation.
- **Length**: the closer an arrow reaches the circle, the better the variable is represented by these two components. A short arrow means the variable lives mostly in other dimensions and should *not* be interpreted on this plane.
- **Position relative to the axes**: an arrow lying along the horizontal axis defines PC1; one lying along the vertical axis defines PC2.

The monetary variables form a tight bundle along the positive horizontal axis. `mobile` and `card_tx` point up; `branch`, `age` and `tenure` point down. `utilization`, `log_loan` and `missed` have short arrows here — they belong to the third component, not to this plane.

```{r}
#| label: var-circle-13
#| fig-cap: "Variables on the plane of components 1 and 3. The credit-strain block, invisible in the previous figure, appears clearly."
#| fig-width: 7
#| fig-height: 6.5
fviz_pca_var(pca, axes = c(1, 3), col.var = "cos2",
             gradient.cols = c("grey70", "steelblue", "red3"),
             repel = TRUE) +
  labs(title = "Variables: PC1 vs PC3")
```

### Quality of Representation: `cos2`

The quality with which variable $j$ is represented by component $k$ is the squared correlation
$$
\cos^2(j, k) = r(x_j, y_k)^2 = a_{jk}^{2}\lambda_k ,
$$
which is exactly the share of variable $j$'s variance recovered by component $k$. Summing over all $p$ components gives $1$; summing over the retained ones gives the quality of representation on the chosen plane.

```{r}
#| label: cos2-table
var_info <- get_pca_var(pca)
round(var_info$cos2[, 1:3], 3)
```

`log_savings` has $\cos^2 = 0.718$ on PC1: nearly three quarters of its variability is captured by the first component alone. `utilization` has $\cos^2 = 0.105$ on PC1 but $0.672$ on PC3 — reading it on the first plane would be a mistake.

### Communalities

The sum of the squared cosines of a variable over the retained components has a name of its own: it is the **communality** of that variable,
$$
h_j^2 = \sum_{c=1}^{k} \cos^2(j, c) = \sum_{c=1}^{k} a_{jc}^{2}\lambda_c ,
$$
where $k$ is the number of components kept. It is the proportion of variable $j$'s variance that the retained components account for, and $1 - h_j^2$ — the part they leave behind — is called the variable's **uniqueness** or *specific variance*.

The name records what the quantity measures: the part of a variable that it holds *in common* with the others, in the sense that the retained components — which are built from all the variables at once — can reproduce it. What is left over, the uniqueness, is the part of the variable that the shared structure does not reach.

Two properties follow directly from the definition:

- Over **all** $p$ components the communality is exactly $1$, since the components form a complete orthonormal basis. Communalities are only informative once some components have been discarded.
- The **average** communality across the $p$ variables equals the proportion of total variance retained. Communalities are therefore a decomposition, variable by variable, of the single global figure reported by the scree plot.

```{r}
#| label: communalities
communality <- rowSums(var_info$cos2[, 1:3])   # three components retained

round(sort(communality), 3)

c(mean_communality = round(mean(communality), 3),
  variance_retained = round(sum(pca$sdev[1:3]^2) / ncol(X), 3))
```

The two figures agree: $0.672$, the $67.2\%$ of total variance carried by the first three components.

Communalities are the natural check on whether a chosen number of components serves *every* variable or only most of them. Here they range from $0.572$ for `log_current` to $0.777$ for `utilization` — a narrow band, so no variable is being badly misrepresented by the three-component solution. A variable with a communality of, say, $0.15$ would be a warning: it is nearly absent from the retained subspace, and any interpretation of the solution that mentions it would be unfounded. The usual responses are to retain another component, or to accept that the variable measures something the others do not share and to handle it separately.

```{r}
#| label: cos2-plot
#| fig-cap: "Quality of representation of each variable on the first three components."
#| fig-width: 7
#| fig-height: 5.5
corrplot(var_info$cos2[, 1:3], is.corr = FALSE,
         method = "color", tl.col = "black",
         addCoef.col = "grey20", number.cex = 0.7,
         cl.pos = "b")
```

### Contributions

Where `cos2` asks *how well is this variable explained by the component*, contribution asks the reverse: *how much of this component is built from that variable*. It is
$$
\text{contrib}(j,k) = 100 \times a_{jk}^{2},
$$
and the contributions of all variables to a given component sum to $100\%$. The dashed reference line marks $100/p$, the contribution a variable would have if all contributed equally.

```{r}
#| label: contrib
#| fig-cap: "Six largest contributions to each of the first three components. The dashed line is the contribution a variable would have under equal weighting (100/13 = 7.7%)."
#| fig-width: 9
#| fig-height: 3.8
library(gridExtra)
g1 <- fviz_contrib(pca, choice = "var", axes = 1, top = 6, fill = "steelblue") +
      labs(title = "PC1")
g2 <- fviz_contrib(pca, choice = "var", axes = 2, top = 6, fill = "steelblue") +
      labs(title = "PC2")
g3 <- fviz_contrib(pca, choice = "var", axes = 3, top = 6, fill = "steelblue") +
      labs(title = "PC3")
grid.arrange(g1, g2, g3, ncol = 3)
```

:::{.callout-note}
Keep the two summaries apart. **Contributions** are read *down a column* — they say how much each variable builds a given component, and sum to $100\%$ over the variables. **Communalities** are read *along a row* — they say how much of a given variable the retained components recover, and sum to $1$ over all the components.
:::

### Individuals

The scores place each client on the new axes. Colouring them by a variable that did **not** take part in the PCA is one of the most useful checks available: if an external, meaningful grouping lines up with a component, the component is measuring something real.

```{r}
#| label: ind-segment
#| fig-cap: "Clients on the first plane, coloured by commercial segment. The segments separate along PC1, which the PCA discovered without ever seeing them."
#| fig-width: 7.5
#| fig-height: 6
fviz_pca_ind(pca, axes = c(1, 2),
             geom.ind = "point", pointshape = 19, alpha.ind = 0.6,
             col.ind = bank$segment,
             palette = c("grey55", "steelblue", "red3"),
             addEllipses = TRUE, ellipse.type = "norm", ellipse.alpha = 0.08,
             legend.title = "Segment") +
  labs(title = "Clients: PC1 vs PC2")
```

The commercial segment was not among the thirteen variables entered into the PCA, yet the three groups order themselves neatly along PC1:

```{r}
#| label: segment-means
scores <- as.data.frame(pca$x[, 1:3])
agg <- aggregate(scores, by = list(segment = bank$segment), FUN = mean)
agg[, -1] <- round(agg[, -1], 3)
agg
```

Mean PC1 rises from $-1.18$ (Mass) through $1.28$ (Affluent) to $3.25$ (Private), while mean PC2 and PC3 barely move. This is strong external evidence that PC1 is an affluence axis.

### The Biplot

A *biplot* superimposes individuals and variables on the same plane. It is the most informative single figure of a PCA and also the easiest to over-read: the two sets of points are on different scales, and only their *relative* geometry is meaningful.

```{r}
#| label: biplot
#| fig-cap: "Biplot of the bank portfolio. Clients are positioned by their scores, variables by their loadings; a client lying far along an arrow has high values of that variable."
#| fig-width: 8
#| fig-height: 7
fviz_pca_biplot(pca, axes = c(1, 2),
                geom.ind = "point", pointshape = 19,
                alpha.ind = 0.35, col.ind = bank$segment,
                palette = c("grey55", "steelblue", "red3"),
                col.var = "black", repel = TRUE,
                legend.title = "Segment") +
  labs(title = "Biplot: bank client portfolio")
```

:::{.callout-note}
Reading a biplot:

- A client positioned in the direction of an arrow has an above-average value of that variable.
- Clients far from the origin are atypical on the plane shown; clients near the origin are either average, or badly represented in these two dimensions. The two cases are distinguished by looking at the individual `cos2`, not at the plot.
- The angle between two arrows approximates the correlation between the corresponding variables — but only when both arrows are long.
:::

```{webr-r}
#| label: viz-webr
#| autorun: false

library(factoextra)

X <- make_X()
pca <- prcomp(X, scale. = TRUE)

# change axes to c(1, 3) or c(2, 3) and re-run
fviz_pca_var(pca, axes = c(1, 2), col.var = "cos2", repel = TRUE)
```

```{webr-r}
#| label: viz-webr-ind
#| autorun: false

# colour the clients by churn instead of segment
fviz_pca_ind(pca, axes = c(2, 3), geom.ind = "point",
             col.ind = bank$churn, addEllipses = TRUE,
             legend.title = "Churn")
```

:::{.callout-note appearance="minimal"}
**Packages and commands.** All from **factoextra**: `fviz_pca_var` (correlation circle), `fviz_contrib` (contributions), `fviz_pca_ind` (individuals, with `col.ind` and `addEllipses`) and `fviz_pca_biplot` (biplot). `get_pca_var` returns the numbers behind those figures in `$coord`, `$cos2` and `$contrib`. `corrplot(is.corr = FALSE)` displays a cos2 matrix, and `aggregate` summarises scores by an external grouping.
:::

## Interpreting the Components

A principal component has no meaning until someone gives it one. Interpretation proceeds by reading the loadings of a component and asking what the highly weighted variables have in common. Loadings below roughly $0.3$ in absolute value are usually ignored as noise.

```{r}
#| label: loadings-table
round(pca$rotation[, 1:4], 3)
```

**PC1 — Affluence** ($31.2\%$). Large positive loadings on `log_savings` ($0.421$), `log_limit` ($0.418$), `log_income` ($0.414$), `log_invest` ($0.412$) and `log_current` ($0.355$); everything else is small. This is a *size* factor: a client high on PC1 has more of everything monetary. Its negative loadings on `utilization` ($-0.161$) and `missed` ($-0.148$) are weak but sensible — wealthier clients use less of their limit and miss fewer payments.

**PC2 — Digital engagement versus branch and life stage** ($22.6\%$). Positive on `card_tx` ($0.484$) and `mobile` ($0.467$); negative on `branch` ($-0.445$), `age` ($-0.399$) and `tenure` ($-0.393$). This is a *contrast* factor, and it says something the bank would care about: younger, shorter-tenure clients transact digitally, older and long-standing clients come to the branch. Note that PC2 is essentially uncorrelated with wealth — the digital divide in this portfolio runs along age, not income.

**PC3 — Credit strain, reversed** ($13.5\%$). Negative on `utilization` ($-0.619$), `missed` ($-0.548$) and `log_loan` ($-0.525$). A *low* score on PC3 means high utilisation, missed payments and large outstanding debt. Since it is easier to talk about a factor whose high values mean "more of the thing", we can flip its sign:

```{r}
#| label: flip-pc3
pca_flip <- pca
pca_flip$rotation[, 3] <- -pca$rotation[, 3]
pca_flip$x[, 3]        <- -pca$x[, 3]

round(pca_flip$rotation[, 3], 3)
```

After flipping, a high PC3 is a client under credit strain. The eigenvalue, the variance explained and every distance between clients are unchanged; only the label changes.

:::{.callout-warning}
If you flip a component for readability, flip **both** the loadings and the scores, and do it once, at the start. Flipping only one of the two silently breaks the relationship $\mathbf{Y} = \mathbf{Z}\mathbf{A}$ and every downstream result with it.
:::

**PC4 — a warning** ($7.9\%$). The Kaiser criterion retained it ($\lambda_4 = 1.025$), but its loadings are a mixture: negative on `tenure` ($-0.537$), `age` ($-0.479$), `mobile` ($-0.459$) and `card_tx` ($-0.371$), with nothing else prominent. It appears to be a residual contrast between age and digital use *after* PC2 has taken the part of that contrast the two share. That is a statement about the geometry of the solution, not about clients. It is retainable for prediction and not interpretable as a construct — a distinction worth keeping.

| Component | Variance | Name | High score means |
|:----|:------:|:------------------------|:---------------------------------------|
| PC1 | 31.2% | Affluence | High income, savings, investments and credit limit |
| PC2 | 22.6% | Digital engagement | Young, short tenure, transacts by card and app, rarely visits a branch |
| PC3 (flipped) | 13.5% | Credit strain | High credit utilisation, missed payments, large debt |
| PC4 | 7.9% | — | No stable substantive reading |

:::{.callout-tip}
The interpretation above recovers, almost exactly, the four latent factors used to generate the data (`W` affluence, `A` life stage, `D` digital propensity, `R` credit risk) — although `A` and `D` were deliberately entangled and PCA, which is restricted to *orthogonal* directions, splits them across PC2 and PC4 rather than isolating each.

PCA recovers latent structure only up to a rotation. That is a real limitation, but it is also an opening: if the retained subspace is right and only the axes inside it are awkwardly placed, the axes can be turned. @sec-rotation does exactly that, and recovers all four factors — `A` included.
:::

:::{.callout-note appearance="minimal"}
**Packages and commands.** No new tools: interpretation is reading `pca$rotation`, and flipping a component for readability is a minus sign applied to both `$rotation[, k]` and `$x[, k]`.
:::

## Rotating the Components for Interpretation {#sec-rotation}

Three of our four components had a clear reading; the fourth did not. That is a common outcome, and there is a standard response to it: keep the same set of retained components but **turn the axes inside the subspace they span** until the loadings fall into a pattern that is easier to name.

This is a different operation from the rotation of @sec-first-pc. There, rotating the axes *was* the PCA — the whole $p$-dimensional basis was turned so that the first axis caught the most variance. Here we take the $k$ retained components as given and rotate only within them, deliberately giving up the maximum-variance ordering in exchange for readability.

### Simple Structure

The target is what @thurstone1947 called **simple structure**: a loadings matrix in which

- each variable loads strongly on **one** component and near zero on the others,
- each component has a handful of large loadings and many small ones,
- few variables load on several components at once.

A matrix like that almost reads itself. The unrotated one rarely looks like that, because the first component is built to absorb as much variance as it can from *every* variable, which guarantees it has many moderate loadings.

Write $\mathbf{L} = \mathbf{A}_k \boldsymbol{\Lambda}_k^{1/2}$ for the $p \times k$ matrix of loadings on the retained components — the variable-component correlations of @sec-visualising. A rotation replaces it by
$$
\mathbf{L}^{*} = \mathbf{L}\,\mathbf{T},
$$
with $\mathbf{T}$ a $k \times k$ orthogonal matrix chosen to make $\mathbf{L}^{*}$ as close to simple structure as some criterion can measure. Throughout this section $k$ is the *number* of retained components (the `k` of the code) and components are indexed by $c$.

:::{.callout-warning}
**"Loadings" is used for two different matrices.** Earlier in the chapter it meant `pca$rotation`, the eigenvectors, whose columns have unit length. Here it means $\mathbf{A}_k\boldsymbol{\Lambda}_k^{1/2}$, whose entries are correlations and whose columns have squared sum $\lambda_c$. Both are standard and both are called loadings; rotation is defined on the second, because simple structure is a statement about correlations. When you compare a printed table with `pca$rotation`, check which one you are holding.
:::

:::{.callout-note}
**What a rotation preserves**

- the subspace spanned by the retained components, so the fitted values and the reconstruction error are unchanged;
- every variable's communality $h_j^2$, and therefore the total variance retained.

**What it gives up**

- the ordering: rotated components are no longer sorted by variance, and none of them is the maximum-variance direction any more;
- the eigenvector property: they are no longer principal components, which is why they are conventionally written RC1, RC2, … rather than PC1, PC2, …;
- uniqueness of the story: $k$ must be chosen *before* rotating, and a different $k$ gives a different rotated solution — not a nested one.
:::

### The Orthomax Family

Most orthogonal rotations maximise one criterion with one parameter changed:
$$
Q(\mathbf{L}^{*}) \;=\; \sum_{c=1}^{k}\sum_{j=1}^{p} l_{jc}^{*4} \;-\; \frac{\gamma}{p}\sum_{c=1}^{k}\left(\sum_{j=1}^{p} l_{jc}^{*2}\right)^{2}.
$$
Different $\gamma$ give the named methods:

| Method | $\gamma$ | Simplifies | Behaviour |
|:-------------|:--------:|:-------------|:-----------------------------------|
| **Quartimax** | $0$ | rows (variables) | Each variable loads on as few components as possible. Tends to leave one large general component. |
| **Varimax** | $1$ | columns (components) | Each component gets a few large loadings. The default choice, and the one that most reliably produces nameable components. |
| **Equamax** | $k/2$ | both | A compromise; spreads variance more evenly across components. Can be unstable. |
| **Parsimax** | $p(k-1)/(p+k-2)$ | both | A further compromise, weighted by the problem's dimensions. |

Varimax [@kaiser1958] is the one to reach for unless you have a reason not to.

`GPArotation` parameterises the same family through the Crawford-Ferguson constant $\kappa = \gamma/p$, which is why equamax appears below as `kappa = k / (2 * ncol(X))` rather than as $\gamma = k/2$. The two are the same rotation.

### Varimax on the Bank Portfolio

We rotate **four** components rather than three. The reason is the one @sec-how-many left unresolved: Kaiser retained four, broken-stick three, and we kept three because the fourth could not be named. Rotation changes that calculation, as we are about to see.

```{r}
#| label: rot-loadings
k <- 4
L <- pca$rotation[, 1:k] %*% diag(pca$sdev[1:k])
dimnames(L) <- list(colnames(X), paste0("PC", 1:k))

round(L, 3)
```

```{r}
#| label: rot-varimax
vm <- varimax(L)
RC <- vm$loadings[, ]
colnames(RC) <- paste0("RC", 1:k)

round(RC, 3)
```

```{r}
#| label: rot-heatmap
#| echo: false
#| fig-cap: "Loadings on four retained components, before and after a varimax rotation. The rotated panel is what simple structure looks like: one strong cell per row."
#| fig-width: 9.5
#| fig-height: 5.4
long <- rbind(
  data.frame(var = rownames(L),  comp = rep(colnames(L), each = nrow(L)),
             val = as.vector(L),  panel = "Unrotated"),
  data.frame(var = rownames(RC), comp = rep(colnames(RC), each = nrow(RC)),
             val = as.vector(RC), panel = "Varimax rotated"))
long$var  <- factor(long$var, levels = rev(colnames(X)))
long$comp <- factor(long$comp, levels = c(colnames(L), colnames(RC)))

ggplot(long, aes(comp, var, fill = val)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%.2f", val)), size = 2.9,
            colour = ifelse(abs(long$val) > 0.55, "white", "grey25")) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "red3",
                       midpoint = 0, limits = c(-1, 1), name = "loading") +
  facet_wrap(~ panel, scales = "free_x") +
  labs(x = NULL, y = NULL,
       title = "Loadings before and after a varimax rotation",
       subtitle = "four retained components; the same subspace, redescribed") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

The rotated solution is far easier to read, and the change is not cosmetic:

- **RC1 — affluence.** Unchanged in substance: income, savings, investments, limit, current balance, all above $0.75$. Age has dropped out of it ($0.45 \to 0.20$).
- **RC2 — digital engagement.** Now almost pure: `card_tx` $0.90$, `mobile` $0.93$, `branch` $-0.68$. Age and tenure, which contributed $-0.68$ and $-0.67$ to the unrotated PC2, have fallen to $-0.24$ and $-0.20$.
- **RC3 — credit strain.** Essentially untouched.
- **RC4 — life stage.** The component that could not be named is now the cleanest of the four: `age` $-0.91$, `tenure` $-0.93$, and nothing else above $0.4$. A high RC4 is a young, recently acquired client.

Rotation has separated the two constructs that the unrotated solution smeared across PC2 and PC4. Now check what it cost:

```{r}
#| label: rot-invariants
rbind(unrotated = round(100 * colSums(L^2)  / ncol(X), 1),
      varimax   = round(100 * colSums(RC^2) / ncol(X), 1))

c(total_unrotated = round(100 * sum(L^2)  / ncol(X), 1),
  total_varimax   = round(100 * sum(RC^2) / ncol(X), 1))

max(abs(rowSums(RC^2) - rowSums(L^2)))     # communalities
```

The total retained variance is $75.1\%$ either way and no communality moves by more than $10^{-15}$. What changed is the *split*: the first component gave up $31.2\% \to 29.1\%$ and the second $22.6\% \to 17.4\%$, while the fourth rose from $7.9\%$ to $15.0\%$ — and RC4 now explains more than RC3, so the components are genuinely no longer in order.

:::{.callout-tip}
This is the honest argument for keeping the fourth component. Unrotated it was a $7.9\%$ residual nobody could name; rotated it is a $15.0\%$ life-stage factor with two loadings above $0.9$. The number of components to retain and the decision to rotate are not independent questions.
:::

### Comparing Methods

```{r}
#| label: rot-family
#| message: false
library(GPArotation)

crit <- function(M) sum(apply(M^2, 2, function(z) mean(z^2) - mean(z)^2))
cplx <- function(M) mean(rowSums(M^2)^2 / rowSums(M^4))   # Hoffman complexity

sols <- list(Unrotated = L,
             Varimax   = varimax(L, normalize = FALSE)$loadings[, ],
             Quartimax = quartimax(L)$loadings[, ],
             Equamax   = cfT(L, kappa = k / (2 * ncol(X)))$loadings[, ])

data.frame(
  criterion  = round(sapply(sols, crit), 4),
  complexity = round(sapply(sols, cplx), 3),
  var_first  = round(sapply(sols, function(M) 100 * sum(M[, 1]^2) / ncol(X)), 1),
  var_fourth = round(sapply(sols, function(M) 100 * sum(M[, 4]^2) / ncol(X)), 1),
  total      = round(sapply(sols, function(M) 100 * sum(M^2) / ncol(X)), 1))
```

The varimax row here is the *unnormalised* variant, so that all four solutions are scored on the same criterion; the rotation used above is the normalised default, which is why its first component carries $29.1\%$ rather than $28.5\%$.

*Complexity* is the average number of components a variable loads on: it falls from $1.51$ unrotated to about $1.21$ under any of the three rotations. On this portfolio the three orthogonal methods give practically the same answer — quartimax leaves slightly more variance on the first component, equamax slightly less, exactly as their definitions predict. **When the three disagree materially, that is a sign the solution is not stable enough to name.**

:::{.callout-warning}
`stats::varimax()` applies **Kaiser normalisation** by default (`normalize = TRUE`): rows are scaled to unit length before rotating and rescaled afterwards, so that variables with small communalities still get a say. `GPArotation` does not normalise by default. The two therefore optimise slightly different things — on this data the raw criterion reaches $0.3434$ without normalisation and $0.3420$ with it. Neither is wrong, but a comparison across packages must hold the setting fixed.
:::

### Oblique Rotations

Orthogonal rotation keeps the components uncorrelated. That is a convenience, not a fact about the world: real constructs are often related, and forcing right angles can prevent simple structure from being reached at all. **Oblique** rotations — *promax* [@hendrickson1964], *oblimin*, *quartimin* — drop the constraint.

```{r}
#| label: rot-promax
pm  <- promax(L)
Phi <- solve(t(pm$rotmat) %*% pm$rotmat)      # component correlation matrix
dimnames(Phi) <- list(paste0("RC", 1:k), paste0("RC", 1:k))

round(pm$loadings[, ], 3)
round(Phi, 3)
```

Promax sharpens the loadings further — `tenure` reaches $-0.997$ on RC4 and `mobile` $0.984$ on RC2 — and it reports something an orthogonal rotation cannot: the components are correlated, and one pair substantially so. RC2 (digital) and RC4 (life stage) correlate at $0.43$.

That number is not noise. The data were generated with `D` depending on `A` through the term $-0.45\,\mathrm{A}$, which implies a correlation of about $-0.49$ between digital propensity and life stage in the population, and $-0.45$ in these 600 clients; RC4 loads *negatively* on age, making it a "youth" axis, which flips the sign. The oblique rotation has recovered a feature of the data-generating process that the orthogonal solutions, PCA and varimax alike, were constrained from showing.

:::{.callout-note}
With an oblique rotation, read the **pattern** matrix (unique contribution of each component, printed above) for interpretation and the **structure** matrix (plain variable–component correlations, $\mathbf{L}^{*}\boldsymbol{\Phi}$) for a description of association. They differ precisely because the components overlap, and for the same reason the per-component variance shares no longer add up to the total — shared variance would be counted twice.
:::

### Rotated Scores

The scores rotate with the loadings, and they are what you carry into the next analysis. Since the matrix we rotated was $\mathbf{L} = \mathbf{A}_k\boldsymbol{\Lambda}_k^{1/2}$ rather than the eigenvectors themselves, its partner is the **standardised** score matrix, and the rotation multiplies it on the right:
$$
\mathbf{Y}^{*} = \mathbf{Z}\,\mathbf{A}_k\,\boldsymbol{\Lambda}_k^{-1/2},
\qquad
\mathbf{Y}^{*}_{\text{rot}} = \mathbf{Y}^{*}\mathbf{T}.
$$
For an oblique rotation the same holds with $(\mathbf{T}^{\top})^{-1}$ in place of $\mathbf{T}$, and the columns come out correlated — their correlation matrix is exactly $\boldsymbol{\Phi}$.

```{r}
#| label: rot-scores
Ystar    <- scale(X) %*% pca$rotation[, 1:k] %*% diag(1 / pca$sdev[1:k])
RCscores <- Ystar %*% vm$rotmat              # varimax
PMscores <- Ystar %*% solve(t(pm$rotmat))    # promax
colnames(RCscores) <- colnames(PMscores) <- paste0("RC", 1:k)

round(apply(RCscores, 2, sd), 3)                          # unit variance
signif(max(abs(cor(RCscores)[upper.tri(diag(k))])), 3)    # still uncorrelated
signif(max(abs(cor(PMscores) - Phi)), 3)                  # promax: exactly Phi
```

Each column now reads as a score on a named construct: RC4 correlates $-0.906$ with age, so it is a youth-and-recency index on a standard-deviation scale. The promax scores carry the $0.43$ between digital engagement and life stage, so anything downstream that assumes uncorrelated inputs has to allow for it.

:::{.callout-warning}
**Two score conventions.** `prcomp()$x` returns scores with variance $\lambda_c$; the ones above have variance $1$ in every column, because they are the partners of correlation-scaled loadings. Neither is wrong, but they are not interchangeable — $k$-means on the first weights the components by $\lambda$, on the second equally. Say which you used.
:::

### When to Rotate, and When Not To

:::{.callout-caution}
**Rotate** when the goal is to name and communicate constructs — a segmentation brief, a questionnaire's underlying dimensions, a set of risk factors for a tariff discussion.

**Do not rotate** when the goal depends on the properties rotation destroys:

- compression, where you want the fewest axes carrying the most variance;
- a single index — a first component used as a wealth or vulnerability score — since after rotation no component is the best single summary;
- any downstream use of the ordering, such as taking "the first two components".

And whichever you do, say so. "The first four components explain 75%" and "the first four varimax-rotated components explain 75%, split 29/17/14/15 rather than 31/23/14/8" describe different objects.
:::

```{webr-r}
#| label: rot-webr
#| autorun: false

X <- make_X()
pca <- prcomp(X, scale. = TRUE)

k <- 4
L <- pca$rotation[, 1:k] %*% diag(pca$sdev[1:k])
dimnames(L) <- list(colnames(X), paste0("PC", 1:k))

# try k = 2, 3, 5 and watch how much the rotated solution changes
print(round(varimax(L)$loadings, 2), cutoff = 0.3)

# and compare with the oblique solution
print(round(promax(L)$loadings, 2), cutoff = 0.3)
```

:::{.callout-note appearance="minimal"}
**Packages and commands.** **stats** provides `varimax` (with `normalize` controlling Kaiser normalisation) and `promax`; both return `$loadings` and `$rotmat`. **GPArotation** supplies the rest of the family — `quartimax`, and `cfT(kappa = )` for equamax — as well as the oblique `oblimin` and `quartimin`. The component correlation matrix is `solve(t(rotmat) %*% rotmat)`, and rotated scores need `solve` and `t` as shown above.
:::

## Practical Issues

### Outliers and Heavy Tails

PCA maximises variance, and variance is not a robust quantity. A single record far from the rest can rotate a component towards itself, producing a "component" that describes one client.

Suppose data-entry errors inflate the savings balance and the investment holdings of client `C0001` to €50 million and €30 million; the true values are €22 020 and €77 560. We introduce the same error twice — once in the untransformed variables, once in the log-transformed ones — and compare.

```{r}
#| label: outlier-setup
Xr <- bank[, num_vars]                 # untransformed
Xr_bad <- Xr
Xr_bad$balance_savings[1] <- 50e6      # true value 22 020
Xr_bad$investments[1]     <- 30e6      # true value 77 560

X_bad <- X                             # log-transformed
X_bad$log_savings[1] <- log(50e6)
X_bad$log_invest[1]  <- log(30e6)

pca_r  <- prcomp(Xr,     scale. = TRUE)
pca_rb <- prcomp(Xr_bad, scale. = TRUE)
pca_lb <- prcomp(X_bad,  scale. = TRUE)

pv <- function(q) round(100 * q$sdev[1:3]^2 / ncol(X), 1)
rbind(untransformed_clean     = pv(pca_r),
      untransformed_corrupted = pv(pca_rb),
      log_clean               = pv(pca),
      log_corrupted           = pv(pca_lb))
```

On the log scale the corrupted record barely registers: the variance explained by the first three components moves from $31.2\%$, $22.6\%$, $13.5\%$ to $30.8\%$, $22.6\%$, $13.5\%$. On the untransformed scale the solution falls apart. Compare the first component before and after:

```{r}
#| label: outlier-loadings
cbind(clean = round(pca_r$rotation[, 1], 3),
      corrupted = round(pca_rb$rotation[, 1], 3))

cor(pca_r$rotation[, 1], pca_rb$rotation[, 1])
```

The correlation between the two versions of PC1 is $-0.02$: the affluence factor has not been perturbed, it has been **displaced entirely**, and the first component of the corrupted analysis is now the age-versus-digital contrast. The affluence structure did not merely move down the ordering — the corrupted variables, having almost all their standardised variance concentrated in one point, are now nearly constant across the other $599$ clients and no longer correlate with anything.

Where did the corrupted client go? Into a component of its own:

```{r}
#| label: outlier-hijack
round(pca_rb$x[1, ] / apply(pca_rb$x, 2, sd), 1)   # its scores, in standard deviations
round(pca_rb$rotation[, 3], 3)                     # the component it created
```

The third component of the corrupted analysis takes $15.3\%$ of the total variance, loads $0.70$ on each of the two corrupted variables and essentially zero on everything else, and the offending client sits $24$ standard deviations along it. That component is not a description of the portfolio; it is a description of one typing mistake.

```{r}
#| label: outlier-demo
#| fig-cap: "Left: the clean untransformed analysis, where PC1 is affluence. Right: the same data with two corrupted cells, on the plane where the damage is visible. The third component of the corrupted analysis exists only to hold one client (red)."
#| fig-width: 9
#| fig-height: 4.5
par(mfrow = c(1, 2))
plot(pca_r$x[, 1], pca_r$x[, 2], pch = 19, col = "grey60", cex = 0.6,
     xlab = "PC1", ylab = "PC2", main = "Clean data: PC1 vs PC2")
plot(pca_rb$x[, 1], pca_rb$x[, 3], pch = 19, col = "grey60", cex = 0.6,
     xlab = "PC1", ylab = "PC3", main = "Corrupted data: PC1 vs PC3")
points(pca_rb$x[1, 1], pca_rb$x[1, 3], pch = 19, col = "red3", cex = 1.6)
par(mfrow = c(1, 1))
```

All of this followed from an error in two cells of a table of $7\,800$ values.

:::{.callout-tip}
Notice what the log transform bought us. It did not remove the outlier — client `C0001` still has an absurd balance — but it compressed errors of a factor of about $2300$ (savings) and $390$ (investments) on the euro scale to additive shifts of $\log(2271) \approx 7.7$ and $\log(387) \approx 6.0$ on the log scale, which are large but not catastrophic. Transformation is a cheap and effective first defence against exactly this failure mode. It is not a substitute for looking at the data.
:::

Practical defences:

- Screen for outliers *before* the PCA using the methods of the Outlier Detection chapter.
- Distinguish errors from genuine extremes. A savings balance of €50 million is an error; a private-banking client with €1.2 million is a real client and should probably stay, possibly after a log transform.
- Use the PCA itself as a diagnostic: an observation with a large score on a *late*, low-variance component, or a large reconstruction error, is atypical in a way the main structure does not explain.
- Consider a robust PCA, based on a robust covariance estimator such as the MCD, when contamination is suspected and cannot be cleaned.

```{r}
#| label: recon-outlier
# reconstruction error per client using three components
Zhat <- pca$x[, 1:3] %*% t(pca$rotation[, 1:3])
err  <- rowSums((scale(X) - Zhat)^2)

head(bank[order(err, decreasing = TRUE), c("client_id", "segment", "utilization",
                                           "missed_payments", "mobile_logins")], 5)
```

These are the clients least well described by the three-component summary — the natural shortlist for an anomaly review.

### Missing Values

`prcomp()` has no mechanism for missing data: it removes any row with an `NA`, which can be a substantial part of the sample when many variables are involved. The options are those of the Missing Values chapter — deletion, single imputation, or multiple imputation — with one addition specific to PCA: iterative PCA imputation, which fills the missing cells with the values predicted by a low-rank reconstruction and iterates until convergence (implemented in the `missMDA` package).

```{r}
#| label: missing-demo
X_na <- X
set.seed(1)
X_na[sample(nrow(X), 60), "log_income"] <- NA

nrow(na.omit(X_na))          # rows surviving listwise deletion
```

Losing $10\%$ of the sample to missingness in *one* variable is already noticeable; with missingness spread over several variables the loss compounds quickly.

### Categorical Variables

PCA is defined for numeric variables. `segment`, `region` and `churn` cannot enter the analysis directly. Three options:

- **Leave them out and use them for illustration**, as we did above — colouring the individuals plot by segment. This is the safest choice and often the most informative.
- **Dummy-code them**. Workable for binary variables, but a dummy has variance $\pi(1-\pi)$, which is small for rare categories; standardising then inflates rare categories and lets them drive components.
- **Use a method designed for the data type**: Multiple Correspondence Analysis (MCA) for a set of categorical variables, or Factor Analysis of Mixed Data (FAMD) when numeric and categorical variables must be analysed together. Both are available in the `FactoMineR` package.

### Multicollinearity and Near-Constant Variables

PCA tolerates perfect collinearity — the corresponding eigenvalue is simply zero — but a variable with zero variance breaks the standardisation, since dividing by $s_j = 0$ is undefined. Check before scaling:

```{r}
#| label: zero-var
which(apply(X, 2, sd) == 0)   # integer(0): no constant columns
```

Also worth avoiding is including a variable twice in different guises (an amount and its logarithm, a total and its components). PCA will faithfully report that they are the same variable, and they will jointly dominate a component for no substantive reason.

### Sample Size

PCA estimates $p(p+1)/2$ covariances from $n$ observations. With $p = 13$ that is $91$ quantities; $n = 600$ is comfortable. Rules of thumb suggest at least $5$–$10$ observations per variable, and more when the correlations are weak. With $n < p$ the correlation matrix is singular and at most $n-1$ eigenvalues are non-zero — PCA still runs, but the later components are pure artefact.

:::{.callout-note appearance="minimal"}
**Packages and commands.** `na.omit` shows the cost of listwise deletion, `apply(X, 2, sd)` finds constant columns, and the per-client reconstruction error is `rowSums((scale(X) - Zhat)^2)`. Named but not used here: **missMDA** for iterative PCA imputation, **FactoMineR** for `MCA` and `FAMD` on categorical and mixed data, and a robust covariance estimator such as the MCD for robust PCA.
:::

<!-- COMMENTED OUT: section "PCA as a Preprocessing Step".
     To restore: delete this comment opener and the closing marker below, then
     put the braces back on the 5 opening code fences (r becomes {r}) so knitr
     evaluates them again.

## PCA as a Preprocessing Step {#sec-pca-preprocessing}

The scores are ordinary variables and can be fed into any subsequent method. Two combinations are common.

### PCA Followed by Clustering

Clustering on principal components instead of raw variables removes redundancy, removes correlation between features (so Euclidean distance treats each axis on equal terms), and reduces noise by discarding the low-variance directions.

```r
#| label: pca-kmeans
set.seed(123)
scores3 <- pca$x[, 1:3]
km <- kmeans(scores3, centers = 3, nstart = 25)

round(km$centers, 2)
table(km$cluster)
```

```r
#| label: pca-kmeans-plot
#| fig-cap: "k-means with k = 3 on the first three component scores, displayed on the PC1-PC2 plane. The axes are the components themselves, not a further reduction."
#| fig-width: 7.5
#| fig-height: 6
clus <- data.frame(scores3, cluster = factor(km$cluster))

ggplot(clus, aes(PC1, PC2, colour = cluster)) +
  geom_point(size = 1.6, alpha = 0.7) +
  stat_ellipse(type = "norm", linewidth = 0.8) +
  scale_colour_manual(values = c("steelblue", "red3", "darkgreen")) +
  labs(title = "k-means on principal component scores",
       x = "PC1 (affluence)", y = "PC2 (digital engagement)") +
  theme_minimal()
```

The clusters have a reading in terms of the named components: one group is affluent and digital, one is affluent and branch-based, one is mass-market. Cross-tabulating against the commercial segment shows how much the behavioural grouping agrees with the bank's existing one:

```r
#| label: cluster-vs-segment
table(cluster = km$cluster, segment = bank$segment)
```

:::{.callout-warning}
Clustering on principal components is a *choice*, not an improvement by default. Discarded components sometimes contain exactly the direction that separates two groups — a small-variance direction can still be the one that matters. Compare the clustering on components with the clustering on the full standardised data before adopting it, and use the validation tools of the Clustering chapter (silhouette, Dunn, gap statistic) to decide.
:::

### PCA Followed by Logistic Regression

Regressing a response on principal components — *principal component regression* — is a standard remedy for multicollinearity. Here we model churn.

```r
#| label: pcr
dat_pc <- data.frame(churn = bank$churn, pca$x[, 1:4])
m_pc   <- glm(churn ~ PC1 + PC2 + PC3 + PC4, data = dat_pc, family = binomial)
round(summary(m_pc)$coefficients, 3)
```

The result is instructive. **PC1 — the component explaining the most variance — is not significant** ($p = 0.64$), while PC2, PC3 and PC4 all are. Affluence, the dominant source of variation in the portfolio, tells us nothing about who leaves; digital disengagement and credit strain do.

```r
#| label: pcr-compare
m_full <- glm(churn ~ ., data = data.frame(churn = bank$churn, X), family = binomial)
c(AIC_4_components = round(AIC(m_pc), 1),
  AIC_13_variables = round(AIC(m_full), 1))
```

Four components fit as well as thirteen variables, with nine fewer parameters.

:::{.callout-warning}
**PCA is unsupervised.** It orders components by variance in the predictors, with no reference to the response. There is no guarantee that the leading components are the useful ones — as PC1 demonstrates here. Selecting components by their eigenvalues and then regressing is therefore a gamble; selecting them by their relationship to the response is legitimate but reintroduces the selection problem PCA was meant to avoid.

When the aim is prediction rather than description, consider *partial least squares* (PLS), which builds components that maximise covariance with the response, or simply a regularised regression (ridge, lasso) on the original variables.
:::

:::{.callout-note appearance="minimal"}
**Packages and commands.** `kmeans(scores, centers = , nstart = )` clusters the score matrix, `table` cross-tabulates the result against a known grouping, and `glm(family = binomial)` with `AIC` fits and compares the regressions on components.
:::

END OF COMMENTED-OUT SECTION -->

## Common Pitfalls

:::{.callout-caution}
1. **Forgetting to scale** when variables have different units — the analysis then reflects the units.
2. **Interpreting a component before checking `cos2`** — a variable near the origin of the correlation circle is not "average", it is *absent* from that plane.
3. **Reading the sign of a component as meaningful** — the sign is arbitrary; only the contrast is real.
4. **Assuming the first component is the important one** — it is only the largest. PCA orders components by variance in the variables, with no reference to any outcome you might later want to model, so importance depends on the question being asked.
5. **Applying PCA to a set of uncorrelated variables** — the eigenvalues will all be near $1$ and nothing is gained. Look at the correlation matrix first.
6. **Computing the PCA on the full data set before splitting into train and test sets** — the test data then influence the transformation, and performance estimates are optimistic. Fit `prcomp()` on the training set and apply it with `predict()`.
7. **Treating principal components as causes** — they are descriptive summaries of covariation, and nothing about the method licenses a causal reading.
8. **Reporting components nobody can name** without saying so — an uninterpretable component is a legitimate result, but it should be labelled as such.
:::

```{r}
#| label: train-test
#| eval: false
# The correct way to use PCA inside a predictive workflow
set.seed(42)
idx        <- sample(nrow(X), 0.7 * nrow(X))
pca_train  <- prcomp(X[idx, ], scale. = TRUE)          # fitted on training data only
scores_tr  <- pca_train$x[, 1:4]
scores_te  <- predict(pca_train, newdata = X[-idx, ])[, 1:4]
```

:::{.callout-note appearance="minimal"}
**Packages and commands.** The train/test pattern is the only code here: fit `prcomp` on the training rows, then apply the *same* transformation to the test rows with `predict(pca_train, newdata = )`. Never refit.
:::

## Beyond PCA: Other Routes to Fewer Dimensions

PCA makes three commitments: the new variables are *linear* combinations of the old ones, the criterion is *global* variance, and the whole $n \times p$ matrix has to be decomposed at once. Each one is relaxed by a family of methods, and on large data the third becomes the binding constraint.

**Multidimensional scaling (MDS)** starts from a matrix of distances rather than from the variables themselves. Classical MDS — also called principal coordinates analysis — applied to Euclidean distances is not a rival to PCA; it *is* PCA:

```{r}
#| label: mds-check
mds <- cmdscale(dist(scale(X)), k = 2)

signif(max(abs(abs(mds) - abs(pca$x[, 1:2]))), 3)   # identical, up to the sign of each axis
```

The interesting case is the one PCA cannot reach. Feed MDS a dissimilarity that is *not* Euclidean — Gower's coefficient for mixed numeric and categorical data, a correlation distance, an edit distance between strings — and you get a map of the clients even though no variable matrix exists. *Non-metric* MDS goes further still, preserving only the rank order of the dissimilarities, which is the right choice when the dissimilarities are ordinal.

**Nonlinear methods** drop the linearity. Kernel PCA runs an ordinary PCA in a feature space reached through a kernel. *t*-SNE and UMAP take a different view again: rather than preserving variance or global distance, they try to keep each point's *near neighbours* near, which is what makes them so effective at showing clusters in high-dimensional data. UMAP is the faster of the two, keeps more of the global structure, and — unlike *t*-SNE — can place new points on an existing map.

**On large data**, the pressure is usually computational rather than conceptual, and it comes in two forms. With many rows and columns, a truncated or randomised SVD returns the leading $k$ components without ever forming the full decomposition, and works on sparse matrices that would not fit in memory as a dense one. With very many columns, random projection is cheaper still: project onto random directions and, by the Johnson–Lindenstrauss lemma, pairwise distances survive to within a controllable error. Autoencoders — neural networks trained to reconstruct their own input through a narrow layer — are the nonlinear version of the same idea, and scale to data far beyond what `prcomp` will hold.

**NIPALS** — *nonlinear iterative partial least squares* [@wold1966] — is the oldest algorithm that computes only the leading components. It never forms $\mathbf{S}$. Each component is found by alternating two least-squares regressions on the residual matrix $\mathbf{E}$, which starts as $\mathbf{Z}$. Given a guess at the scores $\mathbf{y}$, regress each column of $\mathbf{E}$ on $\mathbf{y}$ and normalise the coefficients to obtain the loadings $\mathbf{a}$; then regress each row of $\mathbf{E}$ on $\mathbf{a}$ to update the scores, $\mathbf{y} = \mathbf{E}\mathbf{a}$. Repeat until $\mathbf{y}$ settles, then *deflate*, $\mathbf{E} \leftarrow \mathbf{E} - \mathbf{y}\mathbf{a}^{\top}$, and start on the next component. Combining the two regressions gives $\mathbf{a} \propto \mathbf{E}^{\top}\mathbf{E}\,\mathbf{a}$, so this is the power method, converging to the eigenvector of the largest remaining eigenvalue at a rate set by the ratio of the next eigenvalue to that one. Because each regression is a sum over cells, it can simply skip missing ones, and NIPALS gives a PCA of incomplete data without deleting any client — provided the cells are missing at random.

| Method | Idea | In R |
|:------------------|:------------------------------------|:-------------------------|
| Classical MDS | Coordinates from a distance matrix; Euclidean case = PCA | `cmdscale` (**stats**) |
| Non-metric MDS | Preserve the rank order of dissimilarities | `isoMDS` (**MASS**), `metaMDS` (**vegan**), **smacof** |
| Kernel PCA | PCA in a feature space defined by a kernel | `kpca` (**kernlab**) |
| *t*-SNE | Preserve local neighbourhoods, for visualisation | `Rtsne` (**Rtsne**) |
| UMAP | As above, faster, and can map new points | `umap` (**uwot**) |
| Truncated / randomised SVD | The same leading components, computed cheaply | `prcomp_irlba` (**irlba**), **RSpectra** |
| NIPALS | Leading components one at a time by alternating regressions; tolerates missing cells | `nipals` (**nipals**, **ade4**), `pca(method = "nipals")` (**pcaMethods**) |
| Autoencoder | Nonlinear compression learned by a network | **keras**, **torch** |

:::{.callout-warning}
*t*-SNE and UMAP are for **looking**, not for measuring. They have no loadings, so no component can be named; they are stochastic, so two runs differ; and the distances between clusters, along with the apparent sizes of the clusters, carry no reliable meaning. Both depend strongly on a neighbourhood-size setting (`perplexity`, `n_neighbors`) that can manufacture groups that are not there.

For the kind of tabular client data in this chapter, PCA remains the default — it is deterministic, its components have loadings you can interpret, and `predict` applies the same transformation to new clients. Reach for a neighbour-embedding method when the goal is a picture of a high-dimensional space, and report it as a picture.
:::

:::{.callout-note appearance="minimal"}
**Packages and commands.** `cmdscale` and `dist` are in **stats**; `isoMDS` is in **MASS**, which ships with R. Everything else is an install: **vegan** or **smacof** for non-metric MDS, **kernlab** for `kpca`, **Rtsne** and **uwot** for *t*-SNE and UMAP, **irlba** (`prcomp_irlba`) or **RSpectra** for truncated SVD on large or sparse matrices, **nipals** or **ade4** (`nipals`) and the Bioconductor package **pcaMethods** for NIPALS, **keras** or **torch** for autoencoders.
:::

## Summary

- PCA replaces $p$ correlated variables by $p$ uncorrelated components, ordered by the variance they explain, and is useful only to the extent that the original variables are correlated.
- Check suitability before fitting. Bartlett's test of sphericity must reject, but at realistic sample sizes it nearly always does; the KMO measure ($0.78$ here) and the per-variable MSA say whether the correlation is shared widely enough for a few components to summarise it.
- The components are the eigenvectors of the covariance (or correlation) matrix, and the variance each explains is the corresponding eigenvalue. Equivalently, they come from the SVD of the centred data matrix.
- Centering is mandatory. Scaling is mandatory whenever the variables are in different units, which for client data is nearly always. Log-transform skewed monetary variables first.
- The eigenvalues of a correlation matrix sum to $p$, so a component with $\lambda > 1$ explains more than one original variable.
- Choose the number of components with several criteria at once — scree, cumulative variance, Kaiser, broken stick — and let interpretability settle the disagreements.
- Loadings describe variables, scores describe observations, `cos2` measures how well a variable is represented, contributions measure how much a variable builds a component, and the communality $h_j^2$ — the sum of a variable's squared cosines over the retained components — measures how much of that variable the reduced solution recovers.
- In our bank portfolio, three components explaining $67\%$ of the variance recover an *affluence* axis, a *digital engagement versus branch* axis, and a *credit strain* axis.
- Rotating the retained components — varimax and its relatives — buys interpretability by giving up the variance ordering. It leaves the subspace, the communalities and the total retained variance untouched, and redistributes variance among the components. Here it turns an unnameable fourth component into the clearest factor of the four.
- PCA is unsupervised: the largest component need not be the most useful for any particular downstream task.

## Exercises

### Exercise 1 — Scaling

Run a PCA on the raw (untransformed, unscaled) bank variables and compare it with the PCA on the correlation matrix of the transformed data.

1. What percentage of variance does the first component explain in each case?
2. Which variables have loadings above $0.3$ on PC1 in each case?
3. Convert the six monetary variables (`income`, `balance_current`, `balance_savings`, `investments`, `credit_limit`, `loan_balance`) from euros to *thousands* of euros and re-run both analyses. Which one changes, and why?

```{webr-r}
#| label: ex1
#| autorun: false

num_vars <- c("age", "tenure", "income", "balance_current", "balance_savings",
              "investments", "credit_limit", "utilization", "loan_balance",
              "card_tx", "mobile_logins", "branch_visits", "missed_payments")

pca_raw <- prcomp(bank[, num_vars], scale. = FALSE)

# your turn
```

### Exercise 2 — Reproducing the output by hand

Using only `cor()`, `eigen()` and matrix multiplication:

1. Compute the eigenvalues of the correlation matrix of `X` and verify that they sum to $13$.
2. Compute the scores of the first three components as $\mathbf{Y} = \mathbf{Z}\mathbf{A}$, where $\mathbf{Z}$ is the standardised data.
3. Verify that the score columns have variances equal to the eigenvalues, and that their pairwise correlations are zero.
4. Verify that $\cos^2(j,k) = a_{jk}^2\lambda_k$ equals the squared correlation between variable $j$ and score $k$.

```{webr-r}
#| label: ex2
#| autorun: false

X <- make_X()

Z <- scale(X)
e <- eigen(cor(X))

# your turn
```

:::: {.callout-note appearance="minimal"}
**Hint**: `Y <- Z %*% e$vectors` gives all the scores at once. Compare `apply(Y, 2, var)` with `e$values`, and `round(cor(Y[, 1:3]), 10)` with the identity matrix.
::::

### Exercise 3 — How many components?

For the PCA of `X`:

1. Produce the scree plot and state where you would place the elbow.
2. Apply the Kaiser criterion and the broken-stick model. They disagree — explain the disagreement in terms of $\lambda_4$.
3. How many components are needed to reach $80\%$ of the total variance? Is that a reasonable target here?
4. Examine the loadings of PC5 and PC6. Can you name either of them? Does that change your answer to (1)?

```{webr-r}
#| label: ex3
#| autorun: false

X   <- make_X()
pca <- prcomp(X, scale. = TRUE)

p <- ncol(X)
bstick <- sapply(1:p, function(k) sum(1 / (k:p)) / p) * 100

# your turn
```

### Exercise 4 — An insurance portfolio

Adapt the analysis to an insurance setting. The code below generates a portfolio of $500$ motor policies.

```{webr-r}
#| label: ex4
#| autorun: true

set.seed(7)
m <- 500
V <- rnorm(m)   # vehicle and coverage value
B <- rnorm(m)   # driving behaviour / risk

policy <- data.frame(
  policy_id      = sprintf("P%04d", 1:m),
  driver_age     = round(pmin(pmax(rnorm(m, 46 - 4 * B, 12), 18), 85)),
  licence_years  = NA,
  vehicle_age    = round(pmin(pmax(rnorm(m, 9 - 3 * V, 4), 0), 30)),
  vehicle_value  = round(exp(rnorm(m, log(14000) + 0.55 * V, 0.45)), -2),
  annual_premium = round(exp(rnorm(m, log(520) + 0.40 * V + 0.35 * B, 0.30)), 0),
  coverage_limit = round(exp(rnorm(m, log(50000) + 0.60 * V, 0.35)), -3),
  deductible     = round(exp(rnorm(m, log(400) - 0.30 * B, 0.40)), -1),
  annual_km      = round(exp(rnorm(m, log(12000) + 0.35 * B, 0.40)), -2),
  claims_3y      = rpois(m, exp(log(0.45) + 0.85 * B)),
  claim_cost_3y  = NA,
  bonus_malus    = round(pmin(pmax(rnorm(m, 0.90 + 0.16 * B, 0.12), 0.5), 2.0), 2),
  years_insured  = round(pmin(pmax(rnorm(m, 11 - 2 * B, 5), 0.5), 40), 1)
)
policy$licence_years <- round(pmin(policy$driver_age - 18,
                                   pmax(rnorm(m, policy$driver_age - 24, 4), 0.5)), 1)
policy$claim_cost_3y <- round(policy$claims_3y *
                              exp(rnorm(m, log(2200) + 0.5 * V, 0.6)), -1)

str(policy)
```

1. Which variables should be log-transformed before the PCA? Justify with a skewness calculation.
2. `claim_cost_3y` is zero for policies with no claims. What does that do to a log transform, and how will you handle it?
3. Fit the PCA on the correlation matrix, decide how many components to keep, and name them.
4. Produce the correlation circle and the biplot, colouring policies by `claims_3y > 0`.
5. Compare your components with the two latent variables (`V`, `B`) used to generate the data. Did PCA recover them?
6. Rotate your retained components with varimax. Does the rotated solution separate the vehicle-and-coverage dimension from the driving-behaviour dimension more cleanly than the unrotated one? Compare each solution's correlations with `V` and `B`.

### Exercise 5 — Rotation and interpretation

1. Build the loadings matrix $\mathbf{L}$ for **three** retained components and apply a varimax rotation. Compare it with the four-component rotation of @sec-rotation. Does a clean life-stage factor appear? What does that tell you about the order in which "how many components?" and "rotate or not?" have to be decided?
2. Verify on your own solution that the rotation leaves every communality and the total retained variance unchanged, while changing the share carried by each component.
3. Return to the four-component solution of @sec-rotation (`k <- 4`) and rotate it with quartimax and with equamax as well. Which of the three leaves the most variance on the first component, and why is that exactly what its criterion is built to do?
4. On the same four components, run `varimax()` with `normalize = TRUE` and with `normalize = FALSE`. Report both criterion values and explain which variables Kaiser normalisation is protecting.
5. Apply `promax()` to the four-component solution. Report the component correlation matrix. Which pair of components is most strongly correlated, and does the code that generated `bank` account for it? Repeat with three components: what happens to that correlation, and why?
6. A colleague writes: *"the first rotated component explains 29.1% of the variance, so it is the most important dimension in the portfolio."* Give two reasons to be careful with that sentence.

```{webr-r}
#| label: ex5
#| autorun: false

X   <- make_X()
pca <- prcomp(X, scale. = TRUE)

k <- 3
L <- pca$rotation[, 1:k] %*% diag(pca$sdev[1:k])
dimnames(L) <- list(colnames(X), paste0("PC", 1:k))

varimax(L)

# questions 3-5 use the four-component solution: set k <- 4 and rebuild L

# your turn
```

<!-- COMMENTED OUT: Exercise 6.
     To restore: delete this comment opener and the closing marker at the end of the file.

### Exercise 6 — Components and churn

1. Fit a logistic regression of `churn` on the first four principal components and on the thirteen original variables. Compare AIC and the significance of the coefficients.
2. Which component is most strongly associated with churn? Express what that means in words, using the interpretation of that component.
3. PC1 is not significant. Is that a failure of PCA? Explain.
4. Split the data into $70\%$ training and $30\%$ test. Fit the PCA on the training set only, apply it to the test set with `predict()`, and compare the test-set accuracy of the two models. Why would fitting the PCA on the full data set be wrong?

```{webr-r}
#| label: ex6
#| autorun: false

pca <- prcomp(X, scale. = TRUE)
dat <- data.frame(churn = bank$churn, pca$x[, 1:4])
m_pc <- glm(churn ~ PC1 + PC2 + PC3 + PC4, data = dat, family = binomial)
summary(m_pc)

# your turn
```

END OF COMMENTED-OUT EXERCISE -->

## References {.unnumbered}

::: {#refs}
:::
